---
title: "rquery"
output: github_document
date: "2017-12-04"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

`rquery` is a experimental/demonstration of a simplified sequenced query language based
on [Codd's relational algebra](https://en.wikipedia.org/wiki/Relational_algebra) and not
currently recommended for non-experimental use.  `rquery` is something we whipped up in a singe
weekend to see how small a scope such an adapter might have.  

A few things `rquery` is missing include:

  * `rquery` does not have the table lifetime management tools.
  * Substitution of values from the environment into expressions.
  * Quoting of column names within complex expressions.
  
All of the above are easy to add, but not central to the one-off experiment.
`rquery` is not for production use, but can be an excellent advanced `SQL`
training tool (it shows how some very deep `SQL` by composing `rquery` operators).  Right
now `rquery` is biased towards `PostgeSQL` `SQL`.

There are many other relational algebra inspired specialized query languages.

  * [Alpha](https://en.wikipedia.org/wiki/Alpha_(programming_language))
  * [QUEL](https://en.wikipedia.org/wiki/QUEL_query_languages)
  * [Tutorial D](https://en.wikipedia.org/wiki/D_(data_language_specification)#Tutorial_D)
  * [LINQ](https://msdn.microsoft.com/en-us/library/bb308959.aspx)
  * [SQL](https://en.wikipedia.org/wiki/SQL)
  * (and possibly) [`dplyr`](http://dplyr.tidyverse.org)
 

`rquery`
is a thin translation to `SQL` layer, but we are trying to put the Codd relational operators
front and center (user their naming, and back-port `SQL` progress such as window functions
to the appropriate relational operator).  `rquery` differs from `dplyr` in that `rquery` 
is trying to stay near the Codd relational operators (in particular grouping is a transient
state inside the `rquery::extend()` operator, not a durable user visible annotation as with `dplyr::group_by()`).

The primary relational operators are:

  * `extend()`.  Extend adds derived columns to a relation table.  With a sufficiently powerful `SQL` provider this includes ordered and partitioned window functions.
  * `project()`.  Project is usually portrait as the equivalent to column selection.  In our opinion the original relational nature of the operator is best captured by moving `SQL`'s "`GROUP BY`" aggregation functionality to this operator.
  * `natural_join()`.  This is the relational join operator, using all common columns as the equi-join condition.  The next operator to add would definitely be `theta-join` as that adds a lot more expressiveness to the grammar.
  * `select_rows()`.  This is Codd's relational row selection.  Obviously `select` alone is an over-used and now ambiguous term (it is the "doit" verb in `SQL` and the *column* selector in `dplyr`).
  
The primary non-relational (traditional `SQL`) operators are:

  * `select_columns()`.  This allows choice of columns (central to `SQL`), but is not a relational operator as it can damage row-uniqueness.
  * `order_by()`.  This is a non-relational "user presentation" verb.  Row order is not well-defined in the relational algebra (and also not in most `SQL` implementations).  If used it should be used last in a query (so it is not undone by later operations).
  
The primary missing relational operators are:

  * Direct rename.
  * Union.
  * Direct set difference, anti-join.
  * Theta-join.
  * Division.
  
Primary useful missing operators:

  * Deselect columns.


A great benefit of Codd's relational algebra is it decomposes data transformations into 
a sequence of operators.  `SQL` loses a lot of the original invariants, and over-specifies
how operations are strung together and insisting on a nesting function notation.  `SQL`
also realizes some of the Codd concepts as operators, some as expressions, and some as predicates
(obscuring the uniformity of the original theory).

A lot of 
the grace of the Codd theory can be recovered through the usual trick changing function composition
notation from `g(f(x))` as `x . f() . g()`.

The `rquery` operators are passive.  They don't do anything other than collect a specification
of the desired calculation.  This data structure can then be printed in a friendly fashion, used to 
generate `SQL`, and (in principle) be the representational layer for a higher-order optimizer.

As an acid test we generate a query equivalent to the non-trivial `dplyr` pipeline 
demonstrated in [Let’s Have Some Sympathy For The Part-time R User](http://www.win-vector.com/blog/2017/08/lets-have-some-sympathy-for-the-part-time-r-user/).

First we set up the database and example data:

```{r ex}
library("rquery")
library('RPostgreSQL')

my_db <- DBI::dbConnect(dbDriver("PostgreSQL"), 
                        host = 'localhost',
                        port = 5432,
                        user = 'postgres',
                        password = 'pg')

d <- dbi_copy_to(my_db, 'd',
                 data.frame(
                   subjectID = c(1,                   
                                 1,
                                 2,                   
                                 2),
                   surveyCategory = c(
                     'withdrawal behavior',
                     'positive re-framing',
                     'withdrawal behavior',
                     'positive re-framing'
                   ),
                   assessmentTotal = c(5,                 
                                       2,
                                       3,                  
                                       4),
                   stringsAsFactors = FALSE),
                 temporary = TRUE, overwrite = TRUE)

print(d)

d %.>%
  to_sql(.) %.>%
  DBI::dbGetQuery(my_db, .) %.>%
  knitr::kable(.)

```

Now we write the calculation in terms of our operators (we have not yet bothered
to add the expression capture features to all of the `rquery` operators, so we are currently simulating it using the development version `1.0.2` of [`wrapr`](https://winvector.github.io/wrapr/);  the package itself works with the production release version of `wrapr`).

```{r calc}
scale <- 0.237

dq <- d %.>%
  extend_nse(.,
             probability :=
               exp(assessmentTotal * scale)/
               sum(exp(assessmentTotal * scale)),
             count := count(1),
             partitionby = 'subjectID') %.>%
  extend_nse(.,
             rank := rank(),
             partitionby = 'subjectID',
             orderby = 'probability')  %.>%
  extend_nse(.,
             isdiagnosis := rank >= count,
             diagnosis := surveyCategory) %.>%
  select_rows_nse(., isdiagnosis) %.>%
  select_columns(., c("subjectID", 
                      "diagnosis", 
                       "probability")) %.>%
  order_by(., 'subjectID')
```

All of the non-standard expression capture (`wrapr::qc()`, `wrapr::qae()`,
substituting in the `scale`, and avoiding the quoted `SQL` column names) could easily
be incorporated into the relation nodes.

We then have our result:

```{r res}
dq %.>%
  to_sql(.) %.>%
  DBI::dbGetQuery(my_db, .) %.>%
  knitr::kable(.)
```

We see we reproduced the result purely in terms of these database operators.

The actual `SQL` query that produces the result is quite involved:

```{r q, comment=''}
cat(to_sql(dq))
```

Part of the hope is the additional record keeping in the operator nodes would 
let a very powerful query optimizer work over the flow before it gets translated
to `SQL`.  At the very least restricting to columns later used and folding selects
together would be achievable.  One should have a good chance at optimization as
the representation is fairly high-level, and many of the operators are relational 
(meaning there are known legal transforms a query optimizer can use).  The flow itself
is represented as follows:

```{r pqp, comment=''}
print(dq)
```

We can even pretty-format it:

```{r pq, comment=''}
cat(gsub("%.>%", "%.>%\n   ", format(dq), fixed = TRUE))
```

And that is our weekend experiment.

```{r cleanup}
DBI::dbDisconnect(my_db)
```


Note: `rquery` is only an experimental package.  Also, it is not currently checked, but
all `rquery` operators should be only used in "zero dependency mode" (never using a value created in the same operator or writing the same value twice) in the sense of [`seplyr::partition_mutate_qt`](https://www.rdocumentation.org/packages/seplyr/versions/0.5.0/topics/partition_mutate_qt).  Again, the point was to see how quickly one can get
a workable data transform pipeline in terms of Codd-inspired operators.


