---
title: "rquery Modes"
output: github_document
---

[`rqdatatable`](https://github.com/WinVector/rqdatatable)/[`rquery`](https://github.com/WinVector/rquery) is designed to have a number of different modes of use.  The primary intended one the considered mode of building up a pipelines from a description of the tables to be acted on.

(Note the `R`/`rquery` version of this example can be found [here](https://github.com/WinVector/rquery/blob/master/Examples/Modes/Modes.md), and the `Python`/`data_algebra` version can be found [here](https://github.com/WinVector/data_algebra/blob/master/Examples/Modes/Modes.md).)

For our example, lets start with the following example data.

```{r}
d <- data.frame(
  x = c(1, 2, 3, 4, 5, 6),
  y = c(2, 2, 2, 3, 7, 10),
  g = c('a', 'a', 'a', 'b', 'b' ,'b'),
  stringsAsFactors = FALSE
)

knitr::kable(d)
```

For our task: let's find a row with the largest ratio of 'y' to 'x', per group 'g'.

The `rquery` concept is to break this into small sub-goals and steps:

 * Find the ratio of 'y' to 'x'.
 * Rank the rows by this ratio.
 * Mark our chosen rows.

In the standard `rquery` practice we build up our processing pipeline to follow our above plan.  The translation involves some familiarity with the `rquery` steps, including the row-numbering command [`row_number()`](https://github.com/WinVector/rquery/blob/master/Examples/WindowFunctions/WindowFunctions.md).

```{r}
library(rqdatatable)
library(rquery)

ops <- local_td(d) %.>%  # Describe table for later operations
  extend(.,       # add a new column
         ratio := y / x) %.>%
  extend(.,       # rank the rows by group and order
         simple_rank := row_number(),
         partitionby = 'g',
         orderby = 'ratio',
         reverse = 'ratio') %.>%
  extend(.,       # mark the rows we want
         choice := simple_rank == 1)
```

The `ops` operator pipeline can than be used to process data.

```{r}
d %.>%
  ops %.>%
  knitr::kable(.)
```

Another point is: this form documents check-able (and enforceable) pre and post conditions on the calculation.  For example such a calculation documents what columns are required by the calculation, and which ones are produced.

```{r}
# columns produced
column_names(ops)

# columns used
columns_used(ops)
```

We can in fact make these conditions the explicit basis of [an interpretation of these data transforms as category theory arrows](https://github.com/WinVector/rquery/blob/master/Examples/Arrow/Arrow.md).

```{r}
arrow(ops)
```

Another way to use `rquery`/`rqdatatable` is in "immediate mode", where we send the data from pipeline stage to pipeline state.

```{r}
d %.>%
  extend(.,       # add a new column
         ratio := y / x) %.>%
  extend(.,       # rank the rows by group and order
         simple_rank := row_number(),
         partitionby = 'g',
         orderby = 'ratio',
         reverse = 'ratio') %.>%
  extend(.,       # mark the rows we want
         choice := simple_rank == 1) %.>%
  knitr::kable(.)
```

Immediate mode skips the `local_td()` step of building a specification for the data to later come, and runs directly on the data.  The advantage of this mode is: it is quick for the user. A disadvantage of this mode is: the pipeline is not left for re-use, and there are possibly expensive data conversions (from `data.frame` to `data.table`) at each operator stage.

```{r}
d %.>%
  wrap %.>%       # wrap data in a description
  extend(.,       # add a new column
         ratio := y / x) %.>%
  extend(.,       # rank the rows by group and order
         simple_rank := row_number(),
         partitionby = 'g',
         orderby = 'ratio',
         reverse = 'ratio') %.>%
  extend(.,       # mark the rows we want
         choice := simple_rank == 1) %.>%
  ex %.>%         # signal construction done, and execute
  knitr::kable(.)
```

The difference is: we use the `wrap` to build a special operator collecting (and checking) pipeline, and then later `ex` to say we are done specifying steps and to apply the operations to the data. Prior to the `ex` step the operator pipeline is available as a field called `underlying` and the set of wrapped `data.frame`s is available as a field called `data_map`.  

Additional execution options are as follows.

We can use `local_td` to convert tables to table descriptions and delay execution.

```{r}
ops <- d %.>%
  local_td %.>%   # capture table description
  extend(.,       # add a new column
         ratio := y / x) %.>%
  extend(.,       # rank the rows by group and order
         simple_rank := row_number(),
         partitionby = 'g',
         orderby = 'ratio',
         reverse = 'ratio') %.>%
  extend(.,       # mark the rows we want
         choice := simple_rank == 1)
```

We can then execute using the pipe.

```{r}
# execute
d %.>%
  ops %.>%
  knitr::kable(.)
```

Or by the `ex_data_table()` method, which is useful for pipelines that contain many tables.

```{r}
ex_data_table(  # useful for multi table pipelines
  ops,
  tables = list('d' = d)) %.>%
  knitr::kable(.)
```



The wrapping of data as a different kind of `rquery` pipeline is an example of using the ["decorator pattern"](https://en.wikipedia.org/wiki/Decorator_pattern) (which can be considered as an object oriented variation of the functional monad pattern, [ref](https://en.wikipedia.org/wiki/Monad_(functional_programming))).  However, these are technical considerations that are the package developer's problem- not problems for the package users. Think of these terms as examples of things developers worry about so users don't have to worry about them.


```{r}
set.seed(2019)
n_rows <- 1000000
d_large <- data.frame(
  x = rnorm(n = n_rows),
  y = rnorm(n = n_rows),
  g = sample(paste0('v_', seq_len(n_rows/10)), 
             size = n_rows, 
             replace = TRUE),
  stringsAsFactors = FALSE
)

f_compiled <- function(dat) {
  dat %.>% ops  # use pre-compiled pipeline
}

f_immediate <- function(dat) {
  dat %.>%
    extend(.,       # add a new column
           ratio := y / x) %.>%
    extend(.,       # rank the rows by group and order
           simple_rank := row_number(),
           partitionby = 'g',
           orderby = 'ratio',
           reverse = 'ratio') %.>%
    extend(.,       # mark the rows we want
           choice := simple_rank == 1)
}

f_wrapped <- function(dat) {
  dat %.>%
    wrap %.>%       # wrap data in a description
    extend(.,       # add a new column
           ratio := y / x) %.>%
    extend(.,       # rank the rows by group and order
           simple_rank := row_number(),
           partitionby = 'g',
           orderby = 'ratio',
           reverse = 'ratio') %.>%
    extend(.,       # mark the rows we want
           choice := simple_rank == 1) %.>%
    ex              # signal construction done, and execute
}
```


And we can also time `data.table` itself (without the translation overhead, though we are adding in the time to convert the `data.frame`).  

```{r, warning=FALSE}
library(data.table)

f_data_table = function(dat) {
  dat <- data.table(dat)
  dat[ , ratio := y / x
       ][order(-ratio) , simple_rank := 1:.N, by = list(g)
          ][ , choice := simple_rank == 1]
}

f_data_table(d) %.>%
  knitr::kable(.)
```


```{r, warning=FALSE}
library(microbenchmark)

timings <- microbenchmark(
  rquery_compiled = f_compiled(d_large),
  rquery_immediate = f_immediate(d_large),
  rquery_wrapped = f_wrapped(d_large),
  data.table = f_data_table(d_large),
  times = 10L
)

print(timings)
```

Notice, the speed differences are usually not that large for short pipelines.  Then intent is: pipeline construction and data conversion steps should be cheap compared to the actual processing steps. 

For those interested: we add `dplyr` and `dtplyr` to the timing comparisons [here](https://github.com/WinVector/rquery/blob/master/Examples/Modes/Modes_dplyr.md).


