---
title: "Modes_dplyr"
output: github_document
---

This is running [`dplyr`](https://CRAN.R-project.org/package=dplyr) and [`dtplyr`]( https://CRAN.R-project.org/package=dtplyr) on the ["`rquery` Modes Example"](https://github.com/WinVector/rquery/blob/master/Examples/Modes/Modes.md).

First we define the functions and data we used in the [`rquery` modes example](https://github.com/WinVector/rquery/blob/master/Examples/Modes/Modes.md) example.

```{r}
library(rquery)
library(data.table)
library(microbenchmark)
```

```{r}
library(dplyr)
packageVersion('dplyr')
library(dtplyr)
packageVersion('dtplyr')
```


```{r}
set.seed(2019)

n_rows <- 1000000

d_large <- data.frame(
  x = rnorm(n = n_rows),
  y = rnorm(n = n_rows),
  g = sample(paste0('v_', seq_len(n_rows/10)), 
             size = n_rows, 
             replace = TRUE),
  stringsAsFactors = FALSE
)
```

```{r}
ops <- local_td(d_large) %.>%  # Describe table for later operations
  extend(.,       # add a new column
         ratio := y / x) %.>%
  extend(.,       # rank the rows by group and order
         simple_rank := row_number(),
         partitionby = 'g',
         orderby = 'ratio',
         reverse = 'ratio') %.>%
  extend(.,       # mark the rows we want
         choice := simple_rank == 1)

d_large %.>%
  ops %.>%
  order_rows(., 'g') %.>%
  select_rows(., choice) %.>% 
  head(.) %.>%
  knitr::kable(.)
```


```{r}
f_compiled <- function(dat) {
  dat %.>% ops  # use pre-compiled pipeline
}

f_immediate <- function(dat) {
  dat %.>%
    extend(.,       # add a new column
           ratio := y / x) %.>%
    extend(.,       # rank the rows by group and order
           simple_rank := row_number(),
           partitionby = 'g',
           orderby = 'ratio',
           reverse = 'ratio') %.>%
    extend(.,       # mark the rows we want
           choice := simple_rank == 1)
}

f_wrapped <- function(dat) {
  dat %.>%
    wrap %.>%       # wrap data in a description
    extend(.,       # add a new column
           ratio := y / x) %.>%
    extend(.,       # rank the rows by group and order
           simple_rank := row_number(),
           partitionby = 'g',
           orderby = 'ratio',
           reverse = 'ratio') %.>%
    extend(.,       # mark the rows we want
           choice := simple_rank == 1) %.>%
    ex              # signal construction done, and execute
}
```

The `dplyr` version of the pipeline is similar, except the window functions are not a single step- but a 4 stage block.

```{r}
f_dplyr <- function(dat) {
  dat %>%
    mutate(        # add a new column
           ratio := y / x) %>%
    group_by(      # rank the rows by group and order
             g) %>%
    arrange( -ratio) %>%
    mutate(     
           simple_rank := row_number()) %>%
    ungroup() %>%  # end of rank block
    mutate(        # mark the rows we want
           choice := simple_rank == 1)
}
```

We are using the most current `CRAN` versions of each (`dtplyr` is currently being re-engineered to try to also cut down the number conversions).

Above we see a key difference between `rquery` and `dplyr`: `rquery` grouped and window functions are single operators in `rquery`, but are driven by annotations between steps in `dplyr`.

`dtplyr` seems to error-out on this problem, meaning the automatic translations from `dplyr` to `data.table` are not sufficient to our task.

```{r, error=TRUE}
f_dplyr(data.table(d_large))
```



We can try to re-write the `dtplyr` pipeline as follows.  It appears switching from `:=` to `=` and replacing `row_number()` with `.I` helps.

```{r}
f_dtplyr <- function(dat) {
  data.table(dat) %>%
    mutate(        # add a new column
           ratio = y / x) %>%
    group_by(      # rank the rows by group and order
             g) %>%
    arrange( -ratio) %>%
    mutate(     
           simple_rank = 1:.N) %>%
    ungroup() %>%  # end of rank block
    mutate(        # mark the rows we want
           choice = simple_rank == 1)
}

res_dtplyr <- f_dtplyr(d_large)

res_dtplyr %.>%
  order_rows(., 'g') %.>%
  select_rows(., choice) %.>% 
  head(.) %.>%
  knitr::kable(.)
```

And we can also time `data.table` itself (without the translation overhead, though we are adding in the time to convert the `data.frame`).

```{r}
f_data_table = function(dat) {
  dat <- data.table(dat)
  dat[ , ratio := y / x
       ][order(-ratio) , simple_rank := 1:.N, by = list(g)
          ][ , choice := simple_rank == 1]
}

res_dt <- f_data_table(d_large)

res_dt %.>%
  order_rows(., 'g') %.>%
  select_rows(., choice) %.>% 
  head(.) %.>%
  knitr::kable(.)
```


```{r}
timings <- microbenchmark(
  rquery_compiled = f_compiled(d_large),
  rquery_immediate = f_immediate(d_large),
  rquery_wrapped = f_wrapped(d_large),
  dplyr = f_dplyr(d_large),
  dtplyr = f_dtplyr(d_large),
  data.table = f_data_table(d_large),
  times = 10L
)

print(timings)
```

For these short pipelines the extra copying in `rquery` immediate mode and `dtplyr` are not causing big problems compared to the overall translation overhead.




