---
title: "Introduction to `rquery`"
output: github_document
---

## Introduction

[`rquery`](https://github.com/WinVector/rquery) is a data wrangling system designed to express complex data manipulation as a series of data transforms. This is in the spirit of `R`'s `base::transform()`, or `dplyr`'s `dplyr::mutate()`.  The operators themselves follow the selections in Codd's relational algebra.  More on the background and context of `rquery` can be found [here](TODO: link to CRAN intro vignette).

In these formulations data manipulation is written as transformations that produce new `data.frame`s, instead of as alterations of a primary data structure (as is the case with `data.table`).  This system uses more space and time. However, in our opinion, it has a number of pedagogical advantages.

In `rquery`'s case the primary set of data operators is as follows:

 * `drop_columns` 
 * `select_columns`
 * `rename_columns`
 * `select_rows`
 * `order_rows`
 * `extend`
 * `project`
 * `natural_join`
 * `convert_records` (supplied by the [`cdata` package](https://github.com/WinVector/cdata)).

These operations break into a small number of themes:

 * Simple column operations (selecting and re-naming columns).
 * Simple row operations (selecting and re-ordering rows).
 * Creating new columns or replacing columns with new calculated values.
 * Aggregating or summarizing data.
 * Combining results between two `data.frame`s.
 * General conversion of record layouts  (supplied by the [`cdata` package]((https://github.com/WinVector/cdata)).
 
The point is: Codd worked out that a great number of data transformations can be decomposed into a small number of the above steps.  `rquery` supplies a high performance implementation of these methods that scales from in-memory scale up through big data scale (to just about anything that supplies a sufficiently powerful `SQL` interface, such as PostgreSQL, Apache Spark, or Google BigQuery).

But lets keep things simple and work through these families of data manipulation operators.

## `rquery` operators

### Simple column operations (selecting and re-naming columns)

The simple column operations are as follows.

 * `drop_columns` 
 * `select_columns`
 * `rename_columns`
 
These operations are easy to demonstrate.

We set up some simple data.

```{r}
d <- data.frame(
  x = c(1, 1, 2),
  y = c(5, 4, 3),
  z = c(6, 7, 8)
)

knitr::kable(d)
```

And the operations are easy to demonstrate.

For example `drop_columns` works as follows, it creates a new `data.frame` without certain columns.

```{r}
library(rquery)

drop_columns(d, c('y', 'z'))
```

In all cases the first argument of a `rquery` operator is either the data to be processed, or an earlier `rquery` pipeline to be extended.  We will take about composing `rquery` operations after we work through examples of all of the basic operations.

We can write the above in piped notation (using the [`wrapr` pipe](https://journal.r-project.org/archive/2018/RJ-2018-042/index.html) in this case):

```{r}
d %.>%
  drop_columns(., c('y', 'z')) %.>%
  knitr::kable(.)
```

Notice the first argument is an explicit "dot" in `wrapr` pipe notation.

`select_columns`'s action is also obvious from example.

```{r}
d %.>%
  select_columns(., c('x', 'y')) %.>%
  knitr::kable(.)
```

### Simple row operations (selecting and re-ordering rows)

The simple row operations are:

 * `select_rows`
 * `order_rows`

`select_rows` keeps the set of rows that meet a given predicate.

```{r}
d %.>%
  select_rows(., x == 1) %.>%
  knitr::kable(.)
```


`order_rows` re-orders rows by a selection of column names (and allows reverse ordering by naming which columns to reverse in the optional `reverse` argument).  Multiple columns can be selected in the order, each column breaking ties in the earlier comparisons.

```{r}
d %.>%
  order_rows(., 
             c('x', 'y'),
             reverse = 'x') %.>%
  knitr::kable(.)
```

General `rquery` operations do not depend on row-order and are not guaranteed to preserve row-order, so if you do want to order rows you should make it the last step of your pipeline.

### Creating new columns or replacing columns with new calculated values

The important create or replace column operation is:

 * `extend`

`extend` accepts arbitrary expressions to create new columns (or replace existing ones).  For example:

```{r}
d %.>%
  extend(., zzz := y / x) %.>%
  knitr::kable(.)
```

We can use `=` or `:=` for column assignment.  In these examples we will use `:=` to keep column assignment clearly distinguishable from argument binding.

`extend` allows for very powerful per-group operations akin to what `SQL` calls ["window functions"](TODO: link).  When the optional `partitionby` argument is set to a vector of column names then aggregate calculations can be performed per-group.  For example.

```{r}
shift <- data.table::shift

d %.>%
  extend(.,
         max_y := max(y),
         shift_z := shift(z),
         row_number := row_number(),
         cumsum_z := cumsum(z),
         partitionby = 'x',
         orderby = c('y', 'z')) %.>%
  knitr::kable(.)
```

Notice the aggregates were performed per-partition (specified by `partitionby`) and in the order determined by the `orderby` argument (without the `orderby` argument order is not guaranteed, so always set `orderby` for windowed operations that depend on row order!).

More on the window functions can be found [here](TODO: link to Window fns tutorial).


### Aggregating or summarizing data

The main aggregation method for `rquery` is:

 * `project`
 
`project` performs per-group calculations, and returns only the grouping columns (specified by `groupby`) and derived aggregates.  For example:

```{r}
d %.>%
  project(.,
         max_y := max(y),
         count := n(),
         groupby = 'x') %.>%
  knitr::kable(.)
```

### Combining results between two `data.frame`s

To combine multiple tables in `rquery` one uses the `natural_join` operator.  In the `rquery` `natural_join` tables are matched by column keys and any two columns with the same name are *coalesced* (meaning the first table with a non-missing values supplies the answer).  This is easiest to demonstrate with an example.

Let's set up new example tables.

```{r}
d_left <- data.frame(
  k = c('a', 'a', 'b'),
  x = c(1, NA, 3),
  y = c(1, NA, NA),
  stringsAsFactors = FALSE
)

knitr::kable(d_left)
```

```{r}
d_right <- data.frame(
  k = c('a', 'b', 'q'),
  y = c(10, 20, 30),
  stringsAsFactors = FALSE
)

knitr::kable(d_right)
```

To perform a join we specify which set of columns our our row-matching conditions (using the `by` argument) and what type of join we want (using the `jointype` argument).  For example we can use `jointype = 'LEFT'` to augment our `d_left` table with additional values from `d_right`.

```{r}
natural_join(d_left, d_right,
             by = 'k',
             jointype = 'LEFT') %.>%
  knitr::kable(.)
```

In a left-join (as above) if the right-table has unique keys then we get a table with the same structure as the left-table- but with more information per row.  This is a very useful type of join in data science projects.  Notice columns with matching names are coalesced into each other, which we interpret as "take the value from the left table, unless it is missing."


### General conversion of record layouts

Record transformation is "simple one you get it".  However, we suggest reading up on that as a separate topic [here](https://github.com/WinVector/cdata).

## Composing operations

We could, of course, perform complicated data manipulation by sequencing `rquery` operations.  For example to select one row with maximal `y` per-`x` group we could work in steps as follows.

```{r}
. <- d
. <- extend(.,
            row_number := row_number(),
            partitionby = 'x',
            orderby = c('y', 'z'))
. <- select_rows(.,
                 row_number == 1)
. <- drop_columns(.,
                  "row_number")
knitr::kable(.)
```

The above discipline has the advantage that it is easy to debug, as we can run line by line and inspect intermediate values.  We can even use the [Bizarrow pipe](TODO: link) to make this look like a pipeline of operations.

```{r}
d ->.;
  extend(.,
         row_number := row_number(),
         partitionby = 'x',
         orderby = c('y', 'z')) ->.;
  select_rows(.,
              row_number == 1) ->.;
  drop_columns(.,
               "row_number") ->.;
  knitr::kable(.)
```

Or we can use the `wrapr` pipe on the data, which we call "immediate mode" (for more on modes please see [here](TODO: link to modes note)).

```{r}
d %.>%
  extend(.,
         row_number := row_number(),
         partitionby = 'x',
         orderby = c('y', 'z')) %.>%
  select_rows(.,
              row_number == 1) %.>%
  drop_columns(.,
               "row_number") %.>%
  knitr::kable(.)
```

However, `rquery` operators can act on `rquery` pipelines in place of data. We can write our operations as follows:

```{r}
ops <- local_td(d) %.>%
  extend(.,
         row_number := row_number(),
         partitionby = 'x',
         orderby = c('y', 'z')) %.>%
  select_rows(.,
              row_number == 1) %.>%
  drop_columns(.,
               "row_number")

cat(format(ops))
```

And we can re-use this pipeline, both on local data and to generate `SQL` to be run in remote databases.

```{r}
d %.>% 
  ops %.>%
  knitr::kable(.)
```

The [`data_algebra`](TODO: link) Python package supplies a nearly identical system for working with data in Python.


## Conclusion

`rquery` supplies a very teachable grammar of data manipulation based on Codd's relational algebra and experience with pipelined data transforms (such as `base::transform()`, `dplyr`, and `data.table`).  

For in-memory situations `rquery` uses `data.table` as it implementation (through the small adapter package `rqdatatable`) and is routinely faster than any other `R` data manipulation system *except* `data.table` itself.

For bigger than memory situations `rquery` can translate to any sufficiently powerful `SQL` dialect, allowing `rquery` pipelines to be executed on PostgeSQL, Apache Spark, or Google BigQuery.


