---
title: "Ad Hoc Queries"
author: "John Mount, Win-Vector LLC"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ad Hoc Queries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`rquery`'s primary workflow is building database operator pipelines that start with
models of the table involved.

Let's try an example. First let's set up our example database and data.

```{r setup}
library("rquery")
db = DBI::dbConnect(RSQLite::SQLite(), 
                    ":memory:")
RSQLite::initExtension(db)

DBI::dbWriteTable(db,
                  'd',
                  data.frame(AUC = 0.6, 
                             R2 = c(0.1, 0.2), 
                             D = NA, z = 2),
                  overwrite = TRUE,
                  temporary = TRUE)
d <- dbi_table(db, 'd')
print(d)

qlook(db, d$table_name)
```

Now we can define a query over this table.

```{r q1}
q <- d %.>%
  select_rows_nse(., R2 > 0.14) %.>%
  extend_nse(., c = sqrt(R2))
```

The idea is:
  * The variable `d` is a table model (name of the table and a set of assumed column names) that allows us to reason about an actual database table to specified later.
  * The query `q` is a sequence of operators we can hold, examine, and alter.

We can print the query/operator pipeline:

```{r q1p, comment=""}
cat(format(q))
```

And we can ask questions of it:

```{r q1q}
column_names(q)

tables_used(q)

columns_used(q)
```

And we can convert the operator pipeline to `SQL` which can then be applied
to an actual database table.

```{r q1s, comment=""}
sql <- to_sql(q, db)
cat(sql)
```

```{r q1e}
DBI::dbGetQuery(db, sql) %.>%
  knitr::kable(.)
```

`rquery` also has an "Ad Hoc" mode for interactive analysis.  
In this mode things are sped up in that the use can work with in-memory tables
and also skip the table modeling step.

Let's first set the global variable `winvector_temp_db_handle` to our 
database handle so the ad hoc mode knows which database to use to implement
the analyses.

```{r defdb}
winvector_temp_db_handle <- list(db = db)
```


We can now run operators directly on in-memory `data.frame`s.

```{r df1}
dL <-  data.frame(AUC = 0.6, 
                  R2 = c(0.1, 0.2), 
                  D = NA, z = 2)

dL %.>%
  select_rows_nse(., R2 > 0.14) %.>%
  knitr::kable(.)

dL %.>%
  select_rows_nse(., R2 > 0.14) %.>%
  extend_nse(., c = sqrt(R2)) %.>%
  knitr::kable(.)
```

Using a function wrapper we can also save ad hoc pipelines for later use.

```{r dfahs}
q2 <- . := {
  select_rows_nse(., R2 > 0.14) %.>%
  extend_nse(., c = sqrt(R2))
}

dL %.>% 
  q2 %.>%
  knitr::kable(.)
```

Or we can use a table model based pipeline directly (without needing additional wrapping).

```{r dm}
q3 <- table_source(table_name = 'tmp', 
                   columns = colnames(dL)) %.>%
  select_rows_nse(., R2 > 0.14) %.>%
  extend_nse(., c = sqrt(R2))

dL %.>% 
  q3 %.>%
  knitr::kable(.)
```


```{r cleanup}
DBI::dbDisconnect(winvector_temp_db_handle$db)
winvector_temp_db_handle <- NULL
```
