---
title: "Pipeable SQL"
author: "John Mount"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipeable SQL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

[`rquery`](https://github.com/WinVector/rquery) is a 
query generator for [`R`](https://www.r-project.org).  It is based on [Edgar F. Codd's relational
algebra](https://en.wikipedia.org/wiki/Relational_algebra) plus experience using `SQL` 
and [`dplyr`](https://CRAN.R-project.org/package=dplyr) at big data
scale.  The design represents an attempt to make `SQL` more teachable by
denoting composition by a sequential pipeline notation instead of nested
queries or functions.  The implementation delivers reliable high 
performance data processing on large data systems such as `Spark`
and databases.
Package features include: data processing trees
or pipelines as observable objects (able to report both columns
produced and columns used), optimized `SQL` generation as an explicit
user visible modeling step, convenience methods for applying query
trees to in-memory `data.frame`s, and low direct package dependencies.

# Pipeable `SQL`

`SQL` is a very powerful data processing (or data engineering) grammar.
Data scientists are well advised to learn to work with `SQL`.  

An inessential difficulty in using `SQL` is `SQL` represents
composition of operations by nesting, which can rapidly become
confusing and illegible.  This can be overcome by using a
query composer such as `rquery` (some more query composers
are listed [here](https://github.com/WinVector/rquery/blob/master/README.md)).

Let's set up our environment so we can work with examples.

```{r chpkg} 
run_vignette <- requireNamespace("DBI", quietly = TRUE) && 
  requireNamespace("RSQLite", quietly = TRUE)
```

```{r setup, eval=run_vignette}
library("rquery")
library("wrapr")

# example database connection
db <- DBI::dbConnect(RSQLite::SQLite(),
                     ":memory:")
RSQLite::initExtension(db)

dbopts <- rq_connection_tests(db)
print(dbopts)
options(dbopts)

# copy in example data
rq_copy_to(
  db, 'd',
  data.frame(v = c(1, -5, 3)),
  temporary = FALSE,
  overwrite = TRUE)

# produce a hande to existing table
d <- db_td(db, "d")
```

`d` is a "table description" which is just
the name of a table and the names of expected columns.
`d` does not store data or a database reference (making it
safe to serialize/de-serialize).  All `rquery` operation trees or
pipelines must start either with a table description or a `data.frame`.
We will discuss table 
descriptions later.

Note: in examples we use `rq_copy_to()` to create data.  This is only for the purpose of having
easy portable examples.  With big data the data is usually already in the remote database or
Spark system. The task is almost always to connect and work with this pre-existing remote data
and the method to do this is [`db_td()`](https://winvector.github.io/rquery/reference/db_td.html),
which builds a reference to a remote table given the table name.  The suggested pattern for working with 
remote tables is to get inputs via [`db_td()`](https://winvector.github.io/rquery/reference/db_td.html)
and land remote results with [`materialze()`](https://winvector.github.io/rquery/reference/materialize.html).
To work with local data one can copy data from memory to the database with [`rq_copy_to()`](https://winvector.github.io/rquery/reference/rq_copy_to.html)
and bring back results with [`execute()`](https://winvector.github.io/rquery/reference/execute.html) (though be aware 
operation on remote non-memory data is `rquery`'s primary intent).


For our first example we will introduce a new column and perform
a calculation using this column.  This is achieved
in `SQL` by writing code in one of two styles: defining the 
first new column
twice (once to land the value and once to use), or sequencing
two queries by nesting.  We will demonstrate both methods.

The define the column twice solution looks like the following.

```{r defcoltwice, eval=run_vignette}
DBI::dbGetQuery(db, "
  SELECT
    *,
    ABS(v) AS absv,
    ABS(v) - v AS delta
  FROM
    d
")
```

In `SQL` the column `absv` is not available for calculation in the
same query that it is produced.  

The nested method looks like the following, we produce the column 
`absv` in one query and then wrap that in another query 
to later use the column.  For expressions longer than `ABS(v)` this
is the preferred solution (until one moves to something like common
table expressions).

```{r nestsql, eval=run_vignette}
DBI::dbGetQuery(db, "
  SELECT
    *,
    absv - v AS delta
  FROM (
    SELECT
      *,
      ABS(v) AS absv
    FROM
      d
  ) subtab
")
```

## `sql_node()`

Using `rquery` we can write the `SQL` composition using pipe notation
(where composition is written as `x %.>% f %.>% g` instead of 
`g(f(x))`).  We are going to use [`wrapr`](https://github.com/WinVector/wrapr) 
[dot-pipe](https://winvector.github.io/wrapr/reference/dot_arrow.html) instead of 
the [`magrittr`](https://CRAN.R-project.org/package=magrittr) pipe to pick up 
a neat feature we will use later (all other examples
will work with the `magrittr` pipe).  The "`%.>%`" glyph can
be [bound to a keyboard shortcut](https://github.com/WinVector/addinexamplesWV) for convenience.

The `rquery` realization of the above calculation is as follows:

```{r rquerypipe1, eval=run_vignette}
op_tree <- d %.>%
  sql_node(., "absv" := "ABS(v)") %.>%
  sql_node(., "delta" := "absv - v")
execute(db, op_tree)
```

The above is what we call "piped `SQL`" and represents a major convenience for
users as the details of how to compose the statements are left to the package.
The `sql_node()` is a very powerful node.  We will use it in our first
few examples and move onto more convenient higher level relational nodes.

We can view the `SQL` translation of the operations tree as follows:

```{r printsql, comment="", eval=run_vignette}
cat(to_sql(op_tree, db))
```

Notice the above translations did not add identifier quotes to our
use of "`v`" in "`ABS(v)`".  This is because the `SQL` expression 
is not parsed in `R`.  If we want to identify terms as variables we
can wrap them with `as.name()` or `quote()` to get the quoting (and other variable
oriented features).  The extra `SELECT` step to pull data from the inner
table is used by `rquery` for important column narrowing steps, and 
can actually improve query performance.

```{r rquerypipe11, comment="", eval=run_vignette}
op_tree <- d %.>%
  sql_node(., "absv" := list(list("ABS(", quote(v), ")"))) %.>%
  sql_node(., "delta" := list(list(quote(absv),"-", quote(v))))
cat(to_sql(op_tree, db))
```

The `list(list())` notation is how we say in `R` that we have a 
single element list (i.e. one expression) that is built up as a list
of terms.  The marking notation is cumbersome, but is not needed when
we move on to relation nodes, which are parsed in `R` and can spot
identifiers without additional help.

`op_tree` itself is a an object with its own presentation format:

```{r printoptree, eval=run_vignette}
cat(format(op_tree))
```

The `op_tree` supplies an number of important summaries about the proposed query:

```{r opsummaries, eval=run_vignette}
column_names(op_tree)

tables_used(op_tree)

columns_used(op_tree)
```

### Composing nodes

We can add nodes to an `op_tree` to build larger operator trees (or pipelines).

```{r addop, eval=run_vignette}
op_tree <- op_tree %.>%
  sql_node(., "prod" := "absv * delta")

cat(format(op_tree))
```

And, the `op_tree` record keeping can be used to catch potential errors
early in pipeline construction.  For example if we try to refer to a non-existent
variable when adding an operator we get an thrown exception (note: a `sql_node()`
being added must have its variables marked as above for pre-checking to occur, 
relational nodes will get this checking automatically).

```{r addoperror, error=TRUE, eval=run_vignette}
op_tree <- op_tree %.>%
  sql_node(., "z" := list(list("1 + ", quote(z))))
```

### A non-trivial example

We can express non-trivial operations in `sql_node()`s.  For example 
we can build a node the calculates for each row how many columns
contain `NA`/`NULL` as is demonstrated here.

```{r countna, eval=run_vignette}
# load up example data
d2 <- rq_copy_to(
  db, 'd2',
  data.frame(v1 = c(1, 2, NA, 3),
             v2 = c(NA, "b", NA, "c"),
             v3 = c(NA, NA, 7, 8),
             stringsAsFactors = FALSE))

# look at table
execute(db, d2)

# get list of columns
vars <- column_names(d2)
print(vars)

# build a NA/NULLs per-row counting expression.
# names are "quoted" by wrapping them with as.name().
# constants can be quoted by an additional list wrapping.
expr <- lapply(vars,
               function(vi) {
                 list("+ (CASE WHEN (",
                      as.name(vi),
                      "IS NULL ) THEN 1.0 ELSE 0.0 END)")
               })
expr <- unlist(expr, recursive = FALSE)
expr <- c(list(0.0), expr)
cat(paste(unlist(expr), collapse = " "))

# instantiate the operator node
op_tree_count_null <- d2 %.>%
  sql_node(., "num_missing" := list(expr))
cat(format(op_tree_count_null))

# examine produced SQL
sql <- to_sql(op_tree_count_null, db)
cat(sql)

# execute
execute(db, op_tree_count_null)
```

And, as this is an important capability, this exact functionality
is wrapped in [`count_null_cols()`](https://winvector.github.io/rquery/reference/count_null_cols.html).

```{r countna2, eval=run_vignette}
# whole process wrapped in convenience node
d2 %.>%
  count_null_cols(., vars, "nnull") %.>%
  execute(db, .)
```

## Working with sets of columns

There is a method to apply a parameterized `SQL` expression to a set
of columns.

```{r psql, eval=run_vignette}
# vector of columns we want to work on
colset <- qc(v1, v2, v3)
# build new names we want as results
colterms <- paste0(colset, "_isNA") := colset
map_to_char(colterms)

# build an apply expression to set of columns query 
s_tree <- d2 %.>%
  sql_expr_set(., colterms, 
               "CASE WHEN . IS NULL THEN 1 ELSE 0 END")
cat(to_sql(s_tree, db))
execute(db, s_tree)
```

# `SQL` first

`rquery` is a "`SQL` first" system.  It is designed to create `SQL`
queries and dispatch them to remote systems (`SQLite`, `Spark`, `PostgreSQL`,
`Redshift`, and other databases) for execution.  The [`execute()`](https://winvector.github.io/rquery/reference/execute.html) method
can be used with big data by adding a `table_name` argument (or also by using the
[`materialize()`](https://winvector.github.io/rquery/reference/materialize.html) method) to land results in a remote table instead of pulling
them back to `R`.

The mantra of `SQL`-first is data starts in the database, and stays in the database (i.e., it is too large to 
depend on round-tripping through `R`).  Another important `SQL`-first package is [`cdata`](https://github.com/WinVector/cdata/)
which provides pure `SQL` based implementations of operators that generalize pivot/un-pivot, cast/melt, or spread/gather.

The better the database implementation the better `rquery` will be, both in terms
of performance and in terms of function (such as the availability of `SQL` window functions).


# Ad-hoc mode

As a convenience `rquery` can work with in-memory `data.frame`s by sending them to 
the `SQL` service provider.  This provider defaults to `RSQlite` or can be
set by setting the global option `rquery.rquery_db_executor`.  We demonstrate
this below.

```{r execd, eval=run_vignette}
old_o <- options(list("rquery.rquery_db_executor" = list(db = db)))

data.frame(v = -2:2) %.>%
  execute(., op_tree)
```

When using the `wrapr` dot pipe the above can be abbreviated as:

```{r rwpipe, eval=run_vignette}
data.frame(v = -2:2) %.>% op_tree
```

The above calculation is managed by [`wrapr` dot pipe
`S3` `wrapr_function`](https://github.com/WinVector/wrapr/blob/master/extras/wrapr_pipe.pdf) extensions.

`rquery` operators can be used directly (without any table description nodes)
when working with in-memory `data.frame`s.

```{r adhocops, eval=run_vignette}
data.frame(x = 5) %.>% sql_node(., "z" := "sqrt(x)")
```

The above calculation is triggered by `S3` override
of any of `print()`, `as.data.frame()` and `head()`.  Remote tables need
an `execute()` or `materialize()` step to specify the database connection.


# Table descriptions

`rquery` table descriptions are simple objects that store only 
the name of a table and expected columns.  Any local data or database
table that has at least the set of columns named in the table description
can be used in a given `rquery` pipeline.

The table description "`d`" we have been using in examples was produced
as a result of moving data to a database by 
[`rq_copy_to()`](https://winvector.github.io/rquery/reference/rq_copy_to.html).
However we can also create a description of an existing database table with
[`db_td()`](https://winvector.github.io/rquery/reference/db_td.html) or
even build a description by hand with 
[`mk_td()`](https://winvector.github.io/rquery/reference/mk_td.html).
Also one can build descriptions of local or in-memory `data.frame`s with
[`local_td()`](https://winvector.github.io/rquery/reference/local_td.html).


# Conclusion

`rquery` is new package, but it is already proving to be correct (avoiding [known data processing issues](http://www.win-vector.com/blog/2018/01/advisory-on-multiple-assignment-dplyrmutate-on-databases/)) and [performant](http://www.win-vector.com/blog/2018/01/advisory-on-multiple-assignment-dplyrmutate-on-databases/).
For working with `R` at a big data scale (say using `PostgreSQL` or `Spark`)
`rquery` is the right specialized tool for specifying data manipulation.

# See also

For deeper dives into specific topics, please see also:

  * [`rquery README`](https://winvector.github.io/rquery/index.html)
  * [Join Controller](https://github.com/WinVector/rquery/blob/master/extras/JoinController.md)
  * [Join Dependency Sorting](https://github.com/WinVector/rquery/blob/master/extras/DependencySorting.md)
  * [PerfTest](https://github.com/WinVector/rquery/blob/master/extras/PerfTest.md)
  * [Assignment Partitioner](https://github.com/WinVector/rquery/blob/master/extras/AssigmentPartitioner.md)
  * [DifferentDBs](https://github.com/WinVector/rquery/blob/master/extras/ExtraDBs.md)
  * [`data.table based` implementation](https://github.com/WinVector/rqdatatable)

------------

# Appendix: Always clean up on the way out

```{r cleanup, eval=run_vignette}
options(old_o)
DBI::dbDisconnect(db)
```

