---
title: "rquery Introduction"
author: "John Mount, Win-Vector LLC"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rquery Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

[`rquery`](https://github.com/WinVector/rquery) is a [`SQL`](https://en.wikipedia.org/wiki/SQL) 
query generator for [`R`](https://www.r-project.org).  It is based on [Edgar F. Codd's relational
algebra](https://en.wikipedia.org/wiki/Relational_algebra) plus experience using `SQL` 
and [`dplyr`](https://CRAN.R-project.org/package=dplyr) at big data
scale.  The design represents an attempt to make `SQL` more teachable by
denoting composition a sequential pipeline notation instead of nested
queries or functions.  The implementation delivers reliable high 
performance data processing on large data systems such as `Spark`
and databases.
Package features include: data processing trees
or pipelines as observable objects (able to report both columns
produced and columns used), optimized `SQL` generation as an explicit
user visible modeling step, convenience methods for applying query
trees to in-memory `data.frame`s, and low direct package dependencies.

# Pipeable `SQL`

`SQL` is a very powerful data processing (or data engineering) grammar.
Data scientists are well advised to learn to work with `SQL`.  

An inessential difficulty of using `SQL` is `SQL` represents
composition of operations by nesting, which can rapidly become
confusing and illegible.

Let's set up our environment so we can work with examples.

```{r setup}
library("rquery")

# example database connection
db <- DBI::dbConnect(RSQLite::SQLite(),
                     ":memory:")
RSQLite::initExtension(db)

# example data
d <- dbi_copy_to(
  db, 'd',
  data.frame(v = c(1, -5, 3)),
  temporary = FALSE,
  overwrite = TRUE)
```

`d` is what we call a "table description" which is just
the name of a table and the names of expected columns.
`d` does not store data or a database reference (making it
safe to serialize/de-serialize).  All `rquery` operation trees or
pipelines must start either with a table description or a `data.frame`.
We will discuss table 
descriptions later.

For our first example we will introduce a new column and perform
a calculation using this column.  This is achieved
in `SQL` by writing code in one of two styles: defining the 
first new column
twice (once to land the value and once to use), or sequencing
two queries by nesting.  We will demonstrate both methods.

The define the column twice solution looks like the following.

```{r defcoltwice}
DBI::dbGetQuery(db, "
  SELECT
    *,
    ABS(v) AS absv,
    ABS(v) - v AS delta
  FROM
    d
")
```

In `SQL` the column `absv` is not available for calculation in the
same query that it is produced.  

The nested method looks like the following, we produce the column 
`absv` in one query and then wrap that in another query 
to later use the column.  For expressions longer than `ABS(v)` this
is the preferred solution.

```{r nestsql}
DBI::dbGetQuery(db, "
  SELECT
    *,
    absv - v AS delta
  FROM (
    SELECT
      *,
      ABS(v) AS absv
    FROM
      d
  ) subtab
")
```

## `sql_node()`

Using `rquery` we can write the `SQL` composition using pipe notation
(where composition is written as `x %.>% f %.>% g` instead of 
`g(f(x))`).  We are going to user [`wrapr`](https://github.com/WinVector/wrapr) 
[dot-pipe](https://winvector.github.io/wrapr/reference/grapes-.-greater-than-grapes.html) instead of 
the [`magrittr`](https://CRAN.R-project.org/package=magrittr) pipe to pick up 
a neat feature we will use later (with the exception of that feature all examples
will work with the `magrittr` pipe).

The `rquery` realization of the above calculation is as follows:

```{r rquerypipe1}
op_tree <- d %.>%
  sql_node(., "absv" := "ABS(v)") %.>%
  sql_node(., "delta" := "absv - v")
execute(db, op_tree)
```

The above is what we call "piped `SQL`" and represents a major convenience for
users as the details of how to compose the statements are left to the package.

We can get the `SQL` equivalent to the operations tree as follows:

```{r printsql, comment=""}
cat(to_sql(op_tree, db))
```

Notice the above expressions did not add identifier quotes to our
use of "`v`" in "`ABS(v)`".  This is because the `SQL` expression 
is not parsed in `R`.  If we want to identify terms as variables we
can wrap them with `as.name()` or `quote()` to get the quoting (and other variable
oriented features).  The extra `SELECT` step to pull data from the inner
table is used by `rquery` for important column narrowing steps, and 
can actually improve query performance.

```{r rquerypipe11, comment=""}
op_tree <- d %.>%
  sql_node(., "absv" := list(list("ABS(", quote(v), ")"))) %.>%
  sql_node(., "delta" := list(list(quote(absv),"-", quote(v))))
cat(to_sql(op_tree, db))
```

The `list(list())` notation is how we say in `R` that we have a 
single element list (i.e. one expression) that is built up as a list
of terms.

`op_tree` is a an object with its own presentation format:

```{r printoptree}
print(op_tree)
```

And `op_tree` stores an number of important summaries about the proposed query:

```{r opsummaries}
column_names(op_tree)

tables_used(op_tree)

columns_used(op_tree)
```

### Composing nodes

We can add nodes to an `op_tree` to build larger operator trees (or pipelines).

```{r addop}
op_tree <- op_tree %.>%
  sql_node(., "prod" := "absv * delta")

cat(format(op_tree))
```

And, the `op_tree` record keeping can be used to catch potential errors
early in pipeline construction.  For example if we try to refer to a non-existent
variable when adding an operator we get an thrown exception (note the node
being added must have its variables marked as above for pre-checking to occur).

```{r addoperror, error=TRUE}
op_tree <- op_tree %.>%
  sql_node(., "z" := list(list("1 + ", quote(z))))
```

# `SQL` first

`rquery` is a "`SQL` first" system.  It is designed only to create `SQL`
queries and dispatch them to remote systems (`SQLite`, `Spark`, `PostgreSQL`,
`Redshift`, and other databases) for execution.  The [`execute()`](https://winvector.github.io/rquery/reference/execute.html) method
can be used with big data by adding a `table_name` argument (or also by using the
[`materialize()`](https://winvector.github.io/rquery/reference/materialize.html) method) to land results in a remote table instead of pulling
them back to `R`.

The better the database implementation the better `rquery` will be, both in terms
of performance and in terms of function (such as the availability of `SQL` window functions).

# Ad-hoc mode

As a convenience `rquery` can work with in-memory `data.frame`s by sending them to 
the `SQL` service provider.  This provider defaults to `RSQlite` or can be
set by setting the global variable `winvector_temp_db_handle`.  We demonstrate
this below.

```{r execd}
winvector_temp_db_handle <- list(db = db)

data.frame(v = -2:2) %.>%
  execute(., op_tree)
```

When using the `wrapr` dot pipe the above can be abbreviated as:

```{r rwpipe}
data.frame(v = -2:2) %.>% op_tree
```

(Note: the above calculation is triggered by `wrapr` dot pipe
`S3` `wrapr_function` definitions.)

Also `rquery` operators can be used directly (without any table description nodes)
when working with in-memory `data.frame`s.

```{r adhocops}
data.frame(x = 5) %.>% sql_node(., "z" := "sqrt(x)")
```

(Note: the above calculation is triggered by `S3` override
of any of `print()`, `as.data.frame()` and `head()`).

# Table descriptions

`rquery` table descriptions are simple objects that store only 
the name of a table and expected columns.  Any local data or database
table that has at least the set of columns named in the table description
can be used in a given `rquery` pipeline.

The table description "`d`" we have been using in examples was produced
as a result of moving data to a database by 
[`dbi_copy_to()`](https://winvector.github.io/rquery/reference/dbi_copy_to.html).
However we can also create a description of an existing table with
[`dbi_table()`](https://winvector.github.io/rquery/reference/dbi_table.html) or
even build a description by hand with 
[`table_source()`](https://winvector.github.io/rquery/reference/table_source.html).

# Operators

The [`sql_node()`](https://winvector.github.io/rquery/reference/sql_node.html) 
alone can make
writing, understanding, and maintaining complex data transformations as queries
much easier.  And this node is a good introduction to some of the power of the 
`rquery` package.  However, the primary purpose of `rquery` is to provide
ready-made relational operators to further simplify to the point of rarely
needing to use the `sql_node()` directly.

The primary operators supplied by `rquery` are:

The primary relational operators include:

  * [`extend_nse()`](https://winvector.github.io/rquery/reference/extend_nse.html)/[`extend_se()`](https://winvector.github.io/rquery/reference/extend_se.html).  Extend adds derived columns to a relation table.  With a sufficiently powerful `SQL` provider this includes ordered and partitioned window functions.  This operator also includes built-in [`seplyr`](https://winvector.github.io/seplyr/)-style [assignment partitioning](https://winvector.github.io/seplyr/articles/MutatePartitioner.html).
  * [`project()`](https://winvector.github.io/rquery/reference/project_nse.html).  Project is usually *portrayed* as the equivalent to column selection, though the original definition includes aggregation.  In our opinion the original relational nature of the operator is best captured by moving `SQL`'s "`GROUP BY`" aggregation functionality.
  * [`natural_join()`](https://winvector.github.io/rquery/reference/natural_join.html).  This a specialized relational join operator, using all common columns as an equi-join condition.
  * [`theta_join()`](https://winvector.github.io/rquery/reference/theta_join_nse.html).  This is the relational join operator allowing an arbitrary matching predicate.
  * [`select_rows()`](https://winvector.github.io/rquery/reference/theta_join_nse.html).  This is Codd's relational row selection.  Obviously `select` alone is an over-used and now ambiguous term (for example: it is already used as the "doit" verb in `SQL` and the *column* selector in `dplyr`).
  * [`rename_columns()`](https://winvector.github.io/rquery/reference/rename_columns.html).  This operator renames sets of columns.
  
The primary non-relational (traditional `SQL`) operators are:

  * [`select_columns()`](https://winvector.github.io/rquery/reference/select_columns.html).  This allows choice of columns (central to `SQL`), but is not a relational operator as it can damage row-uniqueness.
  * [`orderby()`](https://winvector.github.io/rquery/reference/orderby.html). Row order is not a concept in the relational algebra (and also not maintained in most `SQL` implementations). This operator is only useful when used with its `limit=` option, or as the last step as data comes out of the relation store and is moved to `R` (where row-order is usually maintained).

The above list (and especially naming) are chosen to first match Codd's relational concepts (`project`, `select`, `rename`, `join`, aggregation), `SQL` naming 
conventions.  Notice this covers the [primary `dplyr` operators](http://dplyr.tidyverse.org) `mutate()` (Codd's `extend`), `select()` (not relational), `filter()` 
(Codd's `select`, represented in `SQL` by "`WHERE`"), `summarise()` (Codd's `project` or aggregate concepts, triggered in `SQL` by "`GROUP BY`"), `arrange()` (not a 
relational concept, implemented in `SQL` by "ORDER BY").  This correspondence is due to Codd's ideas and `SQL` driving data engineering 
thinking for almost the last 50 years (both with and without credit or citation).

With relational operators the user can work fast and work further away from syntactic details.  For example with a database that supplies window 
functions one can quickly work the "logistic scoring by hand" from  
from [Let’s Have Some Sympathy For The Part-time R User](http://www.win-vector.com/blog/2017/08/lets-have-some-sympathy-for-the-part-time-r-user/).
That example worked with `rquery` code that works with both `PostgreSQL` and `Spark` can be found [here](https://github.com/WinVector/rquery/blob/master/README.md).

# Non-`SQL` node

Not all data transform steps can conveniently be written as a single `SQL`
statement. To work around this potential limitation `rquery` supplies a special
type of node called [`non_sql_node()`](https://winvector.github.io/rquery/reference/non_sql_node.html).
`non_sql_node()` is used to implement arbitrary table to table transforms as
`rquery` pipeline steps.  Two important examples of `non_sql_node()` are
[`materialize_node()`](https://winvector.github.io/rquery/reference/materialize_node.html)
and [`rsummary_node()`](https://winvector.github.io/rquery/reference/rsummary_node.html).

`materialize_node()` can be used to land intermediate results into
a table.  This can be used both to save intermediate calculation stages
and to break up the complexity of long data processing pipelines.

`rsummary_node()` builds a table of summary information about another
database table.  The format is each column of the original table 
produces a row of summary information in the result table.  Here
is a simple example.

```{r rsummaryex}
op_tree %.>%
  rsummary_node(.) %.>%
  execute(db, .)
```

Users can add additional 
capabilities by writing their own `non_sql_node()`s.


# Statement partitioning

# ifelse device

# correctness

# performance

## Comparisons

## Column narrowing


------------

# Appendix: Always clean up on the way out

```{r cleanup}
rm(list = "winvector_temp_db_handle")
DBI::dbDisconnect(db)
```

