---
title: "Substitution"
output: github_document
---

[`rquery`](https://github.com/WinVector/rquery) allows [substitution of variable values for column/variable names](https://cran.r-project.org/web/packages/rquery/vignettes/rquery_substitution.html) in several places. This is particularly useful in the [`extend()`](https://winvector.github.io/rquery/reference/extend.html) and [`select_rows()`](https://winvector.github.io/rquery/reference/select_rows.html) operations.

The substitution is performed using `bquote` "`.()`" notation.  The values being substituted must be of class `name`, to specify they are refering to column names (and not string constants).

Here is an example.

```{r}
d <- data.frame(
  x = c(1, 1, 2),
  y = c(5, 4, 3),
  z = c(6, 7, 8)
)

knitr::kable(d)
```

```{r}
library("rquery")

condition_variable <- as.name('x')
new_value_variable <- as.name('y')
old_value_variable <- as.name('z')

ops <- local_td(d) %.>%
  select_rows(.,
              .(condition_variable) == 1) %.>%
  extend(.,
         .(new_value_variable) := .(old_value_variable) + 1)

cat(format(ops))
```

```{r}
d %.>% 
  ops %.>%
  knitr::kable(.)
```

This also works in the `wrap`/`ex` pattern.

```{r}
d %.>%
  wrap %.>%
  select_rows(.,
              .(condition_variable) == 1) %.>%
  extend(.,
         .(new_value_variable) := .(old_value_variable) + 1) %.>%
  ex %.>%
  knitr::kable(.)
```

And the method can also be used with project.

```{r}
d %.>%
  wrap %.>%
  select_rows(.,
              .(condition_variable) == 1) %.>%
  project(.,
         .(new_value_variable) := max(.(old_value_variable))) %.>%
  ex %.>%
  knitr::kable(.)
```



Note: `rquery` `1.3.9` has an issue with `bquote`/`.()` substitution in "imediate mode".  We have fixed that in later versions.

```{r}
d %.>%
  select_rows(.,
              .(condition_variable) == 1) %.>%
  extend(.,
         .(new_value_variable) := .(old_value_variable) + 1) %.>%
  knitr::kable(.)
```

```{r}
d %.>%
  select_rows(.,
              .(condition_variable) == 1) %.>%
  project(.,
         .(new_value_variable) := max(.(old_value_variable))) %.>%
  knitr::kable(.)
```


