---
title: "column selection"
output: github_document
---

`dplyr` is inconsistent as to which column is selected unless one uses extra notation such as `!!`, `{{}}`, `.data[[]]`, and so on.  Of course if using a name or string directly are not the "correct" notation, why are they allowed?  Notice how different columns are selected in each example, depending on the columns present in the `data.frame`.  The issue is `dplyr` does not commit to an unambiguous interpretation of the basic notation (only the more complicated, longer notations have reliable semantics).

```{r}
library("dplyr")

y = "x"

data.frame(x = 1) %>%
  select(y)

data.frame(x = 1, y = 2) %>%
  select(y)
```

`dplyr` notations that are unambiguous include:

```{r}
data.frame(x = 1) %>%
  select({{y}})

data.frame(x = 1, y = 2) %>%
  select

data.frame(x = 1) %>%
  select(!!y)

data.frame(x = 1, y = 2) %>%
  select(!!y)

data.frame(x = 1) %>%
  select(!!rlang::enquo(y))

data.frame(x = 1, y = 2) %>%
  select(!!rlang::enquo(y))

data.frame(x = 1) %>%
  select(.data[[y]])

data.frame(x = 1, y = 2) %>%
  select(.data[[y]])
```

But other notations don't work (`.data` is apparently a mapping from column names to column indices, and not in fact a reference to the incoming `data.frame`).

```{r, error=TRUE}
data.frame(x = 1) %>%
  select(.data[y])

data.frame(x = 1, y = 2) %>%
  select(.data[y])
```



`R` itself does not have this problem.  Notice how the column named by `y` (which turns out to be `x`) is reliably chosen in all cases.  In `[]` and `[[]]` notations columns are always values (not taken from code or variable names; and `$` always take from code and not from values).

```{r}
y = "x"

data.frame(x = 1)[y]

data.frame(x = 1, y = 2)[y]
```

`rqdatable` also has reliable column selection semantics, columns are always values (not taken from code or variable names).

```{r}
library("rqdatatable")

y = "x"

data.frame(x = 1) %.>% 
  select_columns(., y)

data.frame(x = 1, y = 2) %.>% 
  select_columns(., y)
```
