---
title: "QTiming3"
author: "Win-Vector LLC"
date: "1/11/2018"
output: github_document
---

Let's time [`rquery`](https://winvector.github.io/rquery/), [`dplyr`]( https://CRAN.R-project.org/package=dplyr), and [`data.table`](https://CRAN.R-project.org/package=data.table ) on a non-trivial example.

These timings are on an Amazon m4.2xlarge (8 vCPU, 32GB RAM) running Ubuntu Server 16.04 LTS (HVM),with the database being PostgreSQL 9.6.5 on a db.4.2xlarge instance.

First let's load our
packages, establish a database connection, and declare an [`rquery` ad hoc execution service](https://winvector.github.io/rquery/articles/AdHocQueries.html) (the "`winvector_temp_db_handle`").

```{r dbsetup}
library("rquery")
library("dplyr")
library("microbenchmark")
library("ggplot2")

db <- DBI::dbConnect(RPostgres::Postgres(),
                     host = 'wvdbtest.cv9xgzsc2qa1.us-west-2.rds.amazonaws.com',
                     port = 5432,
                     user = 'wvdbinstance',
                     password = 'wvdb9327262643664',
                     dbname = 'wvdb')

winvector_temp_db_handle <- list(db = db)

packageVersion("rquery")
packageVersion("dplyr")
packageVersion("dbplyr")
packageVersion("DBI")
packageVersion("data.table")
packageVersion("RPostgres")

print(db)
DBI::dbGetQuery(db, "SELECT version()")

R.Version()
```

We now build and extended version of the example from [Let’s Have Some Sympathy For The Part-time R User](http://www.win-vector.com/blog/2017/08/lets-have-some-sympathy-for-the-part-time-r-user/).

```{r data}
nrep <- 10


mkData <- function(nrep) {
  dLocal <- data.frame(
    subjectID = c(1,                   
                  1,
                  2,                   
                  2),
    surveyCategory = c(
      'withdrawal behavior',
      'positive re-framing',
      'withdrawal behavior',
      'positive re-framing'
    ),
    assessmentTotal = c(5,                 
                        2,
                        3,                  
                        4),
    stringsAsFactors = FALSE)
  norig <- nrow(dLocal)
  dLocal <- dLocal[rep(seq_len(norig), nrep), , drop=FALSE]
  dLocal$subjectID <- paste((seq_len(nrow(dLocal)) -1)%/% norig,
                            dLocal$subjectID, 
                            sep = "_")
  rownames(dLocal) <- NULL

  dLocal
}

dLocal <- mkData(nrep)
dR <- rquery::dbi_copy_to(db, 'dR',
                          dLocal,
                          temporary = TRUE, 
                          overwrite = TRUE)
dTbl <- dplyr::tbl(db, dR$table_name)

head(dLocal)

cdata::qlook(db, dR$table_name)

dplyr::glimpse(dTbl)
```

Now we declare our operation pipelines, both on local (in-memory `data.frame`) and
remote (already in a database) data.

```{r query}
scale <- 0.237

# base-R function
# could also try base::split() or base:table()
base_r_calculate_tabular <- function(d) {
  d <- d[order(d$subjectID, d$surveyCategory), , drop=FALSE]
  # compute un-normalized probability
  d$probability <- exp(d$assessmentTotal * scale)
  # set up of for selection
  dmax <- stats::aggregate(d$probability, 
                           by = list(subjectID = d$subjectID), 
                           FUN = max)
  maxv <- dmax$x
  names(maxv) <- dmax$subjectID
  # set up for normalization
  dsum <- stats::aggregate(d$probability, 
                           by = list(subjectID = d$subjectID), 
                           FUN = sum)
  sumv <- dsum$x
  names(sumv) <- dsum$subjectID
  # start selection
  d$maxv <- maxv[d$subjectID]
  d <- d[d$probability >= d$maxv, 
                   , 
                   drop=FALSE]
  # de-dup
  d$rownum <- seq_len(nrow(d))
  drow <-  stats::aggregate(d$rownum, 
                           by = list(subjectID = d$subjectID), 
                           FUN = min)
  minv <- drow$x
  names(minv) <- drow$subjectID
  d$rmin <- minv[d$subjectID]
  d <- d[d$rownum <= d$rmin, , drop=FALSE]
  # renormalize
  d$probability <- d$probability/sumv[d$subjectID]
  d <- d[, c("subjectID", "surveyCategory", "probability")]
  colnames(d)[[2]] <- "diagnosis"
  d
}

# base-R function
# could also try base::split() or base:table()
base_r_calculate_sequenced <- function(d) {
  cats <- base::sort(base::unique(d$surveyCategory))
  res <- NULL
  for(ci in cats) {
    di <- d[d$surveyCategory == ci, , drop=FALSE]
    di <- di[base::order(di$subjectID), , drop=FALSE]
    di$probability <- exp(di$assessmentTotal * scale)
    if(length(base::unique(di$subjectID))!=nrow(di)) {
      stop("base_r_calculate repeated subjectID")
    }
    if(is.null(res)) {
      res <- data.frame(subjectID = di$subjectID,
                        totalProb = di$probability,
                        bestScore = di$probability,
                        diagnosis = ci,
                        stringsAsFactors = FALSE)
    } else {
      if((nrow(di)!=nrow(res)) ||
         (!all(di$subjectID == res$subjectID))) {
        stop("base_r_calculate saw irregular data")
      }
      change <- di$probability > res$bestScore
      res$diagnosis[change] <- ci
      res$bestScore <- base::pmax(res$bestScore, 
                                  di$probability)
      res$totalProb <- res$totalProb + di$probability
    }
  }
  res$probability <- res$bestScore/res$totalProb
  res <- res[, c("subjectID", 
                 "diagnosis", 
                 "probability")]
  res
}



# base-R function
# could also try base::split() or base:table()
base_r_calculate_rows <- function(d) {
  d <- d[base::order(d$subjectID, d$surveyCategory), , drop=FALSE]
  d$probability <- exp(d$assessmentTotal * scale)
  ids <- sort(unique(d$subjectID))
  totals <- numeric(length(ids))
  names(totals) <- ids
  n <- nrow(d)
  choices <- logical(n)
  sum <- 0
  maxID <- 1
  for(i in seq_len(n)) {
    id <- d$subjectID[[i]]
    probi <- d$probability[[i]]
    sum <- sum + probi
    if(probi>d$probability[[maxID]]) {
      maxID <- i
    }
    end_of_group <- (i>=n) || (d$subjectID[[i+1]]!=id)
    if(end_of_group) {
      choices[[maxID]] <- TRUE
      totals[id] <- sum
      sum <- 0
      maxID <- i+1
    }
  }
  d <- d[choices, , drop=FALSE]
  d$probability <- d$probability/totals[d$subjectID]
  d <- d[, c("subjectID", 
               "surveyCategory", 
               "probability")]
  colnames(d)[[2]] <- "diagnosis"
  d
}

# this is a function, 
# so body not evaluated until used
rquery_pipeline <- . := {
  extend_nse(.,
             probability :=
               exp(assessmentTotal * scale)/
               sum(exp(assessmentTotal * scale)),
             count := count(1),
             partitionby = 'subjectID') %.>%
    extend_nse(.,
               rank := rank(),
               partitionby = 'subjectID',
               orderby = c('probability', 'surveyCategory'))  %.>%
    rename_columns(., 'diagnosis' := 'surveyCategory') %.>%
    select_rows_nse(., rank == count) %.>%
    select_columns(., c('subjectID', 
                        'diagnosis', 
                        'probability')) %.>%
    orderby(., 'subjectID') 
}


base_R_row_calculation <- function() {
  base_r_calculate_rows(dLocal)
}

base_R_sequential_calculation <- function() {
  base_r_calculate_sequenced(dLocal)
}

base_R_tabular_calculation <- function() {
  base_r_calculate_tabular(dLocal)
}

rquery_local <- function() {
  dLocal %.>% 
    rquery_pipeline(.) %.>%
    as.data.frame(.) # force execution
}

rquery_database_pull <- function() {
  dR %.>% 
    rquery_pipeline(.) %.>% 
    to_sql(., db) %.>% 
    DBI::dbGetQuery(db, .) %.>%
    as.data.frame(.) # shouldn't be needed
}

rquery_database_land <- function() {
  tabName <- "rquery_tmpx"
  sqlc <- dR %.>% 
    rquery_pipeline(.) %.>% 
    to_sql(., db)
  DBI::dbExecute(db, paste("CREATE TABLE", tabName, "AS", sqlc))
  DBI::dbExecute(db, paste("DROP TABLE", tabName))
  NULL
}

rquery_database_count <- function() {
  dR %.>% 
    rquery_pipeline(.) %.>% 
    sql_node(., "n" := "COUNT(1)") %.>% 
    to_sql(., db) %.>% 
    DBI::dbGetQuery(db, .) %.>%
    as.data.frame(.) # shouldn't be needed
}

# this is a function, 
# so body not evaluated until used
dplyr_pipeline <- . %>%
  group_by(subjectID) %>%
  mutate(probability =
           exp(assessmentTotal * scale)/
           sum(exp(assessmentTotal * scale), na.rm = TRUE)) %>%
  arrange(probability, surveyCategory) %>%
  filter(row_number() == n()) %>%
  ungroup() %>%
  rename(diagnosis = surveyCategory) %>%
  select(subjectID, diagnosis, probability) %>%
  arrange(subjectID)

# this is a function, 
# so body not evaluated until used
# pipeline re-factored to have filter outside
# mutate 
# work around: https://github.com/tidyverse/dplyr/issues/3294
dplyr_pipeline2 <- . %>%
  group_by(subjectID) %>%
  mutate(probability =
           exp(assessmentTotal * scale)/
           sum(exp(assessmentTotal * scale), na.rm = TRUE)) %>%
  arrange(probability, surveyCategory) %>%
  mutate(count = n(), rank = row_number()) %>%
  ungroup() %>%
  filter(count == rank) %>%
  rename(diagnosis = surveyCategory) %>%
  select(subjectID, diagnosis, probability) %>%
  arrange(subjectID)


dplyr_local <- function() {
  dLocal %>% 
    dplyr_pipeline
}

dplyr_local_no_grouped_filter <- function() {
  dLocal %>% 
    dplyr_pipeline2
}

dplyr_tbl <- function() {
  dLocal %>%
    as_tibble %>%
    dplyr_pipeline
}

dplyr_round_trip <- function() {
  dTmp <- dplyr::copy_to(db, dLocal, "dplyr_tmp",
                         # overwrite = TRUE,
                         temporary = TRUE
  )
  res <- dTmp %>% 
    dplyr_pipeline %>%
    collect()
  dplyr::db_drop_table(db, "dplyr_tmp")
  res
}

dplyr_database_pull <- function() {
  dTbl %>% 
    dplyr_pipeline %>%
    collect()
}

dplyr_database_land <- function() {
  tabName = "dplyr_ctmpx"
  dTbl %>% 
    dplyr_pipeline %>%
    compute(name = tabName)
  dplyr::db_drop_table(db, table = tabName)
  NULL
}

dplyr_database_count <- function() {
  dTbl %>% 
    dplyr_pipeline %>%
    tally() %>%
    collect()
}

.datatable.aware <- TRUE

data.table_local <- function() {
  dDT <- data.table::data.table(dLocal)
  dDT[
    , one := 1 ][
      , probability := exp ( assessmentTotal * scale ) / 
        sum ( exp ( assessmentTotal * scale ) ) ,subjectID ][
          , count := sum ( one ) ,subjectID ][
            , rank := rank ( probability ) ,subjectID ][
              rank == count ][
                , diagnosis := surveyCategory ][
                  , c('subjectID', 'diagnosis', 'probability') ][
                    order(subjectID) ]
}

```

Let's inspect the functions.

```{r show}
head(base_R_sequential_calculation())

head(base_R_row_calculation())

head(base_R_tabular_calculation())

head(rquery_local())

rquery_database_land()

head(rquery_database_pull())

rquery_database_count()

head(dplyr_local())

head(dplyr_tbl())

head(dplyr_local_no_grouped_filter())

dplyr_database_land()

head(dplyr_database_pull())

dplyr_database_count()

head(dplyr_round_trip())

head(data.table_local())
```

Now let's measure the speeds with `microbenchmark`.


```{r timings}
timings <- NULL

expressions <- list(
    "rquery in memory" = bquote({ nrow(rquery_local())}),
    # "rquery from db to memory" =  bquote({nrow(rquery_database_pull())}),
    # "rquery database count" =  bquote({rquery_database_count()}),
    # "rquery database land" =  bquote({rquery_database_land()}),
    # "dplyr in memory" =  bquote({nrow(dplyr_local())}),
    # "dplyr tbl in memory" =  bquote({nrow(dplyr_tbl())}),
    "dplyr in memory no grouped filter" =  bquote({nrow(dplyr_local_no_grouped_filter())}),
    "dplyr from memory to db and back" =  bquote({nrow(dplyr_round_trip())}),
    # "dplyr from db to memory" =  bquote({nrow(dplyr_database_pull())}),
    # "dplyr database count" =  bquote({dplyr_database_count()}),
    # "dplyr database land" =  bquote({dplyr_database_land()}),
    "data.table in memory" =  bquote({nrow(data.table_local())}),
    # "base R row calculation" =  bquote({nrow(base_R_row_calculation())}),
    "base R tabular calculation" =  bquote({nrow(base_R_tabular_calculation())}),
    "base R sequential calculation" =  bquote({nrow(base_R_sequential_calculation())})
)

prune <- FALSE

for(nrep in c(1, 10, 100, 1000, 10000, 100000, 1000000, 10000000)) {
  print(nrep)
  dLocal <- mkData(nrep)
  dR <- rquery::dbi_copy_to(db, 'dR',
                            dLocal,
                            temporary = TRUE, 
                            overwrite = TRUE)
  dTbl <- dplyr::tbl(db, dR$table_name)
  tm <- microbenchmark(
    list = expressions,
    times = 5L
  )
  print(tm)
  print(autoplot(tm))
  tmi <- as.data.frame(tm)
  tmi$data_size <- nrow(dLocal)
  timings <- rbind(timings, tmi)
  if(prune) {
    baddies <- unique(tmi$expr[tmi$time > 10*1e+9])
    for(bi in baddies) {
      expressions[[bi]] <- NULL
    }
    if(length(expressions)<=0) {
      break
    }
  }
}
saveRDS(timings, "qtimings3.RDS")
```

`rquery` appears to be relatively fast.  Some of the time for "`rquery` local" is because `rquery`
doesn't *really* have a local mode, it has to copy the data to the database and back
in that case.


```{r sessioninfo}
sessionInfo()
```

```{r dbcleanup}
winvector_temp_db_handle <- NULL
DBI::dbDisconnect(db)
```


