---
title: "QTiming3"
author: "Win-Vector LLC"
date: "1/13/2018"
output: github_document
---

Let's time [`rquery`](https://winvector.github.io/rquery/), [`dplyr`]( https://CRAN.R-project.org/package=dplyr), and [`data.table`](https://CRAN.R-project.org/package=data.table ) on a non-trivial example.

These timings are on an Amazon EC2 c5.4xlarge 16 vcpu 32GB RAM, 128GB block storage, Ubuntu Server 16.04 LTS (HVM).

First let's load our
packages, establish a database connection, and declare an [`rquery` ad hoc execution service](https://winvector.github.io/rquery/articles/AdHocQueries.html) (the "`winvector_temp_db_handle`").

```{r dbsetup}
library("data.table")  # load first so we can overwrite := with rquery
library("rquery")
library("dplyr")
library("microbenchmark")
library("ggplot2")
source("cscan.R")
source("fns.R")

db <- NULL
db <- DBI::dbConnect(RPostgres::Postgres(),
                     host = 'localhost',
                     port = 5432,
                     user = 'ruser',
                     password = 'ruser')
if(!is.null(db)) {
  winvector_temp_db_handle <- list(db = db)
  print(db)
  DBI::dbGetQuery(db, "SELECT version()", stringsAsFactors = FALSE)
}

packageVersion("rquery")
packageVersion("dplyr")
packageVersion("dbplyr")
packageVersion("DBI")
packageVersion("data.table")
packageVersion("RPostgres")

R.Version()
```

We now build and extended version of the example from [Let’s Have Some Sympathy For The Part-time R User](http://www.win-vector.com/blog/2017/08/lets-have-some-sympathy-for-the-part-time-r-user/).

```{r data}
nrep <- 10

dLocal <- mkData(nrep)
head(dLocal)
dR <- NULL
dTbl <- NULL

if(!is.null(db)) {
  dR <- rquery::dbi_copy_to(db, 'dR',
                            dLocal,
                            temporary = TRUE, 
                            overwrite = TRUE)
  dTbl <- dplyr::tbl(db, dR$table_name)
  
  cdata::qlook(db, dR$table_name)
  
  dplyr::glimpse(dTbl)
}
```

Now we declare our operation pipelines, both on local (in-memory `data.frame`) and
remote (already in a database) data.

```{r query}
scale <- 0.237

base_R_row_calculation <- function() {
  base_r_calculate_rows(dLocal)
}

base_R_sequential_calculation <- function() {
  base_r_calculate_sequenced(dLocal)
}

base_R_cframe_calculation <- function() {
  base_r_calculate_cframe(dLocal)
}


base_R_tabular_calculation <- function() {
  base_r_calculate_tabular(dLocal)
}

rquery_local <- function() {
  dLocal %.>% 
    rquery_pipeline(.) %.>%
    as.data.frame(., stringsAsFactors = FALSE) # force execution
}

rquery_database_pull <- function() {
  dR %.>% 
    rquery_pipeline(.) %.>% 
    to_sql(., db) %.>% 
    DBI::dbGetQuery(db, ., stringsAsFactors = FALSE) %.>%
    as.data.frame(., stringsAsFactors = FALSE) # shouldn't be needed
}

rquery_database_land <- function() {
  tabName <- "rquery_tmpx"
  sqlc <- dR %.>% 
    rquery_pipeline(.) %.>% 
    to_sql(., db)
  DBI::dbExecute(db, paste("CREATE TABLE", tabName, "AS", sqlc))
  DBI::dbExecute(db, paste("DROP TABLE", tabName))
  NULL
}

rquery_database_count <- function() {
  dR %.>% 
    rquery_pipeline(.) %.>% 
    sql_node(., "n" := "COUNT(1)") %.>% 
    to_sql(., db) %.>% 
    DBI::dbGetQuery(db, ., stringsAsFactors = FALSE) %.>%
    as.data.frame(., stringsAsFactors = FALSE) # shouldn't be needed
}


dplyr_local <- function() {
  dLocal %>% 
    dplyr_pipeline
}

dplyr_local_no_grouped_filter <- function() {
  dLocal %>% 
    dplyr_pipeline2
}

dplyr_tbl <- function() {
  dLocal %>%
    as_tibble %>%
    dplyr_pipeline
}

dplyr_round_trip <- function() {
  dTmp <- dplyr::copy_to(db, dLocal, "dplyr_tmp",
                         # overwrite = TRUE,
                         temporary = TRUE
  )
  res <- dTmp %>% 
    dplyr_pipeline %>%
    collect()
  dplyr::db_drop_table(db, "dplyr_tmp")
  res
}

dplyr_database_pull <- function() {
  dTbl %>% 
    dplyr_pipeline %>%
    collect()
}

dplyr_database_land <- function() {
  tabName = "dplyr_ctmpx"
  dTbl %>% 
    dplyr_pipeline %>%
    compute(name = tabName)
  dplyr::db_drop_table(db, table = tabName)
  NULL
}

dplyr_database_count <- function() {
  dTbl %>% 
    dplyr_pipeline %>%
    tally() %>%
    collect()
}

```

Let's inspect the functions.

```{r show}

check <- base_R_sequential_calculation()
head(check)

if(!equiv_res(check, base_R_cframe_calculation())) {
  stop("mismatch")
}

if(!equiv_res(check, base_R_row_calculation())) {
  stop("mismatch")
}

if(!equiv_res(check, base_R_tabular_calculation())) {
  stop("mismatch")
}

if(!equiv_res(check, dplyr_local())) {
  stop("mismatch")
}

if(!equiv_res(check, dplyr_tbl())) {
  stop("mismatch")
}

if(!equiv_res(check, dplyr_local_no_grouped_filter())) {
  stop("mismatch")
}

if(!equiv_res(check, data.table_local())) {
  stop("mismatch")
}

if(!is.null(db)) {
  head(rquery_local())
  
  rquery_database_land()
  
  head(rquery_database_pull())
  
  rquery_database_count()
  
  dplyr_database_land()
  
  head(dplyr_database_pull())
  
  dplyr_database_count()
  
  head(dplyr_round_trip())
}
```


Now let's measure the speeds with `microbenchmark`.


```{r timings}
timings <- NULL


expressions <- list(
    # "rquery in memory" = bquote({ nrow(rquery_local())}),
    # "rquery from db to memory" =  bquote({nrow(rquery_database_pull())}),
    # "rquery database count" =  bquote({rquery_database_count()}),
    # "rquery database land" =  bquote({rquery_database_land()}),
    # "dplyr in memory" =  bquote({nrow(dplyr_local())}),
    # "dplyr tbl in memory" =  bquote({nrow(dplyr_tbl())}),
    "dplyr in memory no grouped filter" =  bquote({nrow(dplyr_local_no_grouped_filter())}),
    # "dplyr from memory to db and back" =  bquote({nrow(dplyr_round_trip())}),
    # "dplyr from db to memory" =  bquote({nrow(dplyr_database_pull())}),
    # "dplyr database count" =  bquote({dplyr_database_count()}),
    # "dplyr database land" =  bquote({dplyr_database_land()}),
    "data.table in memory" =  bquote({nrow(data.table_local())}),
    # "base R row calculation" =  bquote({nrow(base_R_row_calculation())}),
    "base R tabular calculation" =  bquote({nrow(base_R_tabular_calculation())}),
    # "base R sequential calculation" =  bquote({nrow(base_R_sequential_calculation())}),
    "base R cframe calculation" =  bquote({nrow(base_R_cframe_calculation())})
)

if(!is.null(db)) {
  expressions <- 
    c(expressions,
      list(
        "rquery from memory to db and back" = bquote({ nrow(rquery_local())}),
        # "rquery from db to memory" =  bquote({nrow(rquery_database_pull())}),
        "rquery database count" =  bquote({rquery_database_count()}),
        "rquery database land" =  bquote({rquery_database_land()}),
        # "dplyr in memory" =  bquote({nrow(dplyr_local())}),
        # "dplyr tbl in memory" =  bquote({nrow(dplyr_tbl())}),
        "dplyr from memory to db and back" =  bquote({nrow(dplyr_round_trip())}),
        # "dplyr from db to memory" =  bquote({nrow(dplyr_database_pull())}),
        "dplyr database count" =  bquote({dplyr_database_count()}),
        "dplyr database land" =  bquote({dplyr_database_land()})
      ))
}

prune <- FALSE

for(nrep in c(1,
              10, 
              100, 
              1000,
              10000,
              100000, 
              1000000)) {
  print(nrep)
  dLocal <- mkData(nrep)
  dR <- NULL
  dTbl <- NULL
  if(!is.null(db)) {
    dR <- rquery::dbi_copy_to(db, 'dR',
                              dLocal,
                              temporary = TRUE, 
                              overwrite = TRUE)
    dTbl <- dplyr::tbl(db, dR$table_name)
  }
  tm <- microbenchmark(
    list = expressions,
    times = 5L
  )
  print(tm)
  print(autoplot(tm))
  tmi <- as.data.frame(tm, stringsAsFactors = FALSE)
  tmi$data_size <- nrow(dLocal)
  timings <- rbind(timings, tmi)
  if(prune) {
    baddies <- unique(tmi$expr[tmi$time > 10*1e+9])
    for(bi in baddies) {
      expressions[[bi]] <- NULL
    }
    if(length(expressions)<=0) {
      break
    }
  }
}
saveRDS(timings, "qtimings3.RDS")
```


```{r sessioninfo}
sessionInfo()
```

```{r dbcleanup}
winvector_temp_db_handle <- NULL
if(!is.null(db)) {
  DBI::dbDisconnect(db)
  db <- NULL
}
```


