---
title: "data.table backend for rquery"
author: "John Mount, Win-Vector LLC"
date: "05/30/2018"
output: github_document
---

We can work an example similar to the [`rquery`](https://winvector.github.io/rquery/) [example](https://winvector.github.io/rquery/index.html) using a [`data.table`](http://r-datatable.com/) 
back-end ([`qdatatable`](https://github.com/WinVector/qdatatable)).

```{r}
library("ggplot2")
library("microbenchmark")
library("dplyr")
library("dtplyr")
# https://github.com/WinVector/qdatatable
library("qdatatable") # devtools::install.packages("WinVector/qdatatable")
```


```{r}
# data example
dL <- build_frame(
   "subjectID", "surveyCategory"     , "assessmentTotal" |
   1          , "withdrawal behavior", 5                 |
   1          , "positive re-framing", 2                 |
   2          , "withdrawal behavior", 3                 |
   2          , "positive re-framing", 4                 )
```


```{r}
scale <- 0.237

# example rquery pipeline
rquery_pipeline <- local_td(dL) %.>%
  extend_nse(.,
             probability :=
               exp(assessmentTotal * scale))  %.>% 
  normalize_cols(.,
                 "probability",
                 partitionby = 'subjectID') %.>%
  pick_top_k(.,
             partitionby = 'subjectID',
             rev_orderby = c('probability', 'surveyCategory')) %.>% 
  rename_columns(., 'diagnosis' := 'surveyCategory') %.>%
  select_columns(., c('subjectID', 
                      'diagnosis', 
                      'probability')) %.>%
  orderby(., cols = 'subjectID')
```

Show expanded form of query tree.

```{r, comment=""}
cat(format(rquery_pipeline))
```

Execute `rquery` pipeline using `data.table` as the impementation.

```{r}
ex_data_table(rquery_pipeline) %.>%
  knitr::kable(.)
```

Execute `rquery` pipeline using `PostgreSQL` as the implementation.

```{r}
# configure a database connection
my_db <- DBI::dbConnect(RPostgreSQL::PostgreSQL(),
                          host = 'localhost',
                          port = 5432,
                          user = 'johnmount',
                          password = '')
dbopts <- rq_connection_tests(my_db)
options(dbopts)
# build the shared handle
winvector_temp_db_handle <- list(db = my_db)


# run the job
execute(dL, rquery_pipeline) %.>%
  knitr::kable(.)
```


`dplyr` pipeline.


```{r}
scale <- 0.237

dplyr_pipeline <- . %>% group_by(subjectID) %>%
  mutate(probability =
           exp(assessmentTotal * scale)/
           sum(exp(assessmentTotal * scale), na.rm = TRUE)) %>%
  arrange(probability, surveyCategory) %>%
  filter(rank() == n()) %>%
  ungroup() %>%
  rename(diagnosis = surveyCategory) %>%
  select(subjectID, diagnosis, probability) %>%
  arrange(subjectID) 

dL %>% 
  dplyr_pipeline %>%
  knitr::kable()
```

Try `dtplyr`.

```{r, error = TRUE}
data.table::as.data.table(dL) %>% 
  dplyr_pipeline
```

Idiomatic `data.table` pipeline.

```{r}
# improved code from:
# http://www.win-vector.com/blog/2018/01/base-r-can-be-fast/#comment-66746
data.table_local <- function(dL) {
  dDT <- data.table::as.data.table(dL)
  dDT <- dDT[, list(diagnosis = surveyCategory,
                    probability = exp (assessmentTotal * scale ) /
                      sum ( exp ( assessmentTotal * scale ) ))
             , subjectID ]
  setorder(dDT, subjectID, probability, -diagnosis)
  dDT <- dDT[, .SD[.N], subjectID]
  setorder(dDT, subjectID)
}

data.table_local(dL) %.>%
  knitr::kable(.)
```


Timings.

```{r}
# fatten up data.frame a bit
dL <- dL[rep(seq_len(nrow(dL)), 100000), , drop = FALSE]
dL$subjectID <- paste(dL$subjectID, (1+seq_len(nrow(dL))) %/% 2, sep = "_")
for(i in seq_len(10)) {
  dL[[paste0("irrelevantCol", i)]] <- runif(nrow(dL))
}

# show we are working on the new larger data
system.time(print(nrow(ex_data_table(rquery_pipeline))))

system.time(print(nrow(dplyr_pipeline(dL))))

system.time(nrow(data.table_local(dL)))
```

```{r}
timings <- microbenchmark(
  nrow(ex_data_table(rquery_pipeline)),
  nrow(data.table_local(dL)),
  nrow(dplyr_pipeline(dL)))
```


```{r}
print(timings)

# summarize by hand using rquery database connector
summary_pipeline <- timings %.>%
  as.data.frame(.) %.>%
  project_nse(., groupby = "expr", mean = avg(time)) 
timings %.>% 
  as.data.frame(.) %.>%
  summary_pipeline %.>%
  knitr::kable(.)

autoplot(timings)

WVPlots::ScatterBoxPlotH(as.data.frame(timings), "time", "expr", "runtime by expression in nanoseconds")
```

For more timings (including fast base-R implementations), please see [here](https://github.com/WinVector/rquery/blob/master/extras/QTimingFollowup/QTiming4.md).

```{r, include=FALSE}
rm(list = "winvector_temp_db_handle")
DBI::dbDisconnect(my_db)
```



