---
title: "Take Care If Trying the RPostgres Package"
author: "Win-Vector LLC"
date: "3/15/2018"
output: github_document
---

Take care if trying the new [`RPostgres`](https://CRAN.R-project.org/package=RPostgres) 
database connection package.  By default it returns some non-standard types that code
developed against other database drivers may not expect, and may not be ready
to defend against.

## Trying the new package

One can try the newer [`RPostgres`](https://CRAN.R-project.org/package=RPostgres)
as a drop-in replacement for the usual [`RPostgreSQL`](https://CRAN.R-project.org/package=RPostgreSQL).

That starts out okay. We can connect to the database 
and and pull a summary about remote data to `R`.


```{r RPostgres_get}
db <- DBI::dbConnect(
  RPostgres::Postgres(),
  host = 'localhost',
  port = 5432,
  user = 'johnmount',
  password = '')
d <- DBI::dbGetQuery(
  db, 
  "SELECT COUNT(1) FROM pg_catalog.pg_tables")
print(d)
ntables <- d$count[[1]]
print(ntables)
```

The result at first looks okay.

```{r RPostgres_examine}
class(ntables)

typeof(ntables)

ntables + 1L

ntables + 1

is.numeric(ntables)
```

But it is only okay, until it is not.

```{r RPostgres_examine2}
pmax(1L, ntables)

pmin(1L, ntables)

ifelse(TRUE, ntables, ntables)

for(ni in ntables) {
  print(ni)
}

unclass(ntables)
```

If your code, or any package code you are using, perform *any* of the above calculations, 
your results will be corrupt and wrong.  It is quite likely any code
written before December 2017 ([`RPostgres`'s first `CRAN` distribution](https://cran.rstudio.com/src/contrib/Archive/RPostgres/))
would not have been written with the `RPostgres` "**`integer64` for all of my friends**" design decision in mind.

Also note, `RPostgres` does not currently appear to write `integer64` back to the database.

```{r writeback}
DBI::dbWriteTable(db, "d", d, 
                  temporary = TRUE, 
                  overwrite = TRUE)
DBI::dbGetQuery(db, "
  SELECT 
     column_name, 
     data_type, 
     numeric_precision, 
     numeric_precision_radix,
     udt_name
  FROM 
     information_schema.columns 
  WHERE 
     table_name = 'd'
  ")
```

```{r RPostgres_disconnect}
DBI::dbDisconnect(db)
```

## The  work-around

The work-around is: add the argument `bigint = "numeric"` to your `dbConnect()` call.  This is mentioned in the [manual](https://cran.r-project.org/web/packages/RPostgres/RPostgres.pdf), but not the default and not called out in
the package description or [`README`](https://cran.r-project.org/web/packages/RPostgres/README.html).  Or, of course,
you could use [`RPostgreSQL`](https://CRAN.R-project.org/package=RPostgreSQL).



