---
title: "QTiming"
author: "Win-Vector LLC"
date: "1/16/2018"
output: github_document
---

Let's time [`rquery`](https://winvector.github.io/rquery/), [`dplyr`]( https://CRAN.R-project.org/package=dplyr), and [`data.table`](https://CRAN.R-project.org/package=data.table ) on a non-trivial example.

These timings are on an late 2014 Mac Mini with 8GB of RAM running OSX 10.12.6, 
version 3.4.3 (2017-11-30) -- "Kite-Eating Tree", and the current (2018-01-07) CRAN versions of all packages (except `rquery`, which is not yet up on CRAN).  We are getting database services from PostgreSQL version `9.6.1` in a docker container.

First let's load our
packages, establish a database connection, and declare an [`rquery` ad hoc execution service](https://winvector.github.io/rquery/articles/AdHocQueries.html) (the "`winvector_temp_db_handle`").

```{r dbsetup}
library("data.table")
library("rquery")
library("dplyr")
library("microbenchmark")
library("ggplot2")

db <- DBI::dbConnect(RPostgres::Postgres(),
                     host = 'localhost',
                     port = 5432,
                     user = 'postgres',
                     password = 'pg')
winvector_temp_db_handle <- list(db = db)

dbopts <- dbi_connection_preferences(db)
print(dbopts)
options(dbopts)

packageVersion("rquery")
packageVersion("dplyr")
packageVersion("dbplyr")
packageVersion("DBI")
packageVersion("data.table")
packageVersion("RPostgres")

print(db)
DBI::dbGetQuery(db, "SELECT version()")

R.Version()
```

We now build and extended version of the example from [Let’s Have Some Sympathy For The Part-time R User](http://www.win-vector.com/blog/2017/08/lets-have-some-sympathy-for-the-part-time-r-user/).

```{r data}
nrep <- 10000

dLocal <- data.frame(
  subjectID = c(1,                   
                1,
                2,                   
                2),
  surveyCategory = c(
    'withdrawal behavior',
    'positive re-framing',
    'withdrawal behavior',
    'positive re-framing'
  ),
  assessmentTotal = c(5,                 
                      2,
                      3,                  
                      4),
  stringsAsFactors = FALSE)
norig <- nrow(dLocal)
dLocal <- dLocal[rep(seq_len(norig), nrep), , drop=FALSE]
dLocal$subjectID <- paste((seq_len(nrow(dLocal)) -1)%/% norig,
                          dLocal$subjectID, 
                          sep = "_")
rownames(dLocal) <- NULL
head(dLocal)

dR <- rquery::dbi_copy_to(db, 'dR',
                          dLocal,
                          temporary = TRUE, 
                          overwrite = TRUE)
cdata::qlook(db, dR$table_name)

dTbl <- dplyr::tbl(db, dR$table_name)
dplyr::glimpse(dTbl)
```

Now we declare our operation pipelines, both on local (in-memory `data.frame`) and
remote (already in a database) data.

```{r query}
scale <- 0.237

# this is a function, 
# so body not evaluated until used
rquery_pipeline <- function(.) {
  extend_nse(.,
             probability :=
               exp(assessmentTotal * scale)/
               sum(exp(assessmentTotal * scale)),
             count := count(1),
             partitionby = 'subjectID') %.>%
    extend_nse(.,
               rank := rank(),
               partitionby = 'subjectID',
               orderby = c('probability', 'surveyCategory'))  %.>%
    rename_columns(., 'diagnosis' := 'surveyCategory') %.>%
    select_rows_nse(., rank == count) %.>%
    select_columns(., c('subjectID', 
                        'diagnosis', 
                        'probability')) %.>%
    orderby(., 'subjectID') 
}


rquery_local <- function() {
  dLocal %.>% 
    rquery_pipeline(.) %.>%
    as.data.frame(.) # force execution
}

rquery_database_pull <- function() {
  dR %.>% 
    rquery_pipeline(.) %.>% 
    to_sql(., db) %.>% 
    DBI::dbGetQuery(db, .) %.>%
    as.data.frame(.) # shouldn't be needed
}

rquery_database_land <- function() {
  tabName <- "rquery_tmpx"
  sqlc <- dR %.>% 
    rquery_pipeline(.) %.>% 
    to_sql(., db)
  DBI::dbExecute(db, paste("CREATE TABLE", tabName, "AS", sqlc))
  DBI::dbExecute(db, paste("DROP TABLE", tabName))
  NULL
}

rquery_database_count <- function() {
  dR %.>% 
    rquery_pipeline(.) %.>% 
    sql_node(., "n" := "COUNT(1)") %.>% 
    to_sql(., db) %.>% 
    DBI::dbGetQuery(db, .) %.>%
    as.data.frame(.) # shouldn't be needed
}

# this is a function, 
# so body not evaluated until used
dplyr_pipeline <- . %>%
  group_by(subjectID) %>%
  mutate(probability =
           exp(assessmentTotal * scale)/
           sum(exp(assessmentTotal * scale), na.rm = TRUE)) %>%
  arrange(probability, surveyCategory) %>%
  filter(row_number() == n()) %>%
  ungroup() %>%
  rename(diagnosis = surveyCategory) %>%
  select(subjectID, diagnosis, probability) %>%
  arrange(subjectID)

# this is a function, 
# so body not evaluated until used
# pipeline re-factored to have filter outside
# mutate 
# work around: https://github.com/tidyverse/dplyr/issues/3294
dplyr_pipeline2 <- . %>%
  group_by(subjectID) %>%
  mutate(probability =
           exp(assessmentTotal * scale)/
           sum(exp(assessmentTotal * scale), na.rm = TRUE)) %>%
  arrange(probability, surveyCategory) %>%
  mutate(count = n(), rank = row_number()) %>%
  ungroup() %>%
  filter(count == rank) %>%
  rename(diagnosis = surveyCategory) %>%
  select(subjectID, diagnosis, probability) %>%
  arrange(subjectID)


dplyr_local <- function() {
  dLocal %>% 
    dplyr_pipeline
}

dplyr_local_no_grouped_filter <- function() {
  dLocal %>% 
    dplyr_pipeline2
}

dplyr_tbl <- function() {
  dLocal %>%
    as_tibble %>%
    dplyr_pipeline
}

dplyr_round_trip <- function() {
  dTmp <- dplyr::copy_to(db, dLocal, "dplyr_tmp",
                         # overwrite = TRUE,
                         temporary = TRUE
  )
  res <- dTmp %>% 
    dplyr_pipeline %>%
    collect()
  dplyr::db_drop_table(db, "dplyr_tmp")
  res
}

dplyr_database_pull <- function() {
  dTbl %>% 
    dplyr_pipeline %>%
    collect()
}

dplyr_database_land <- function() {
  tabName = "dplyr_ctmpx"
  dTbl %>% 
    dplyr_pipeline %>%
    compute(name = tabName)
  dplyr::db_drop_table(db, table = tabName)
  NULL
}

dplyr_database_count <- function() {
  dTbl %>% 
    dplyr_pipeline %>%
    tally() %>%
    collect()
}

.datatable.aware <- TRUE

# improved code from:
# http://www.win-vector.com/blog/2018/01/base-r-can-be-fast/#comment-66746
data.table_local <- function() {
  dDT <- data.table::data.table(dLocal)
  dDT <- dDT[,list(diagnosis = surveyCategory,
                   probability = exp (assessmentTotal * scale ) /
                     sum ( exp ( assessmentTotal * scale ) ))
             ,subjectID ]
  setorder(dDT, subjectID, probability, -diagnosis)
  dDT <- dDT[,.SD[.N],subjectID]
  setorder(dDT, subjectID)
}

```

Let's inspect the functions.

```{r show}
head(rquery_local())

rquery_database_land()

head(rquery_database_pull())

rquery_database_count()

head(dplyr_local())

head(dplyr_tbl())

head(dplyr_local_no_grouped_filter())

dplyr_database_land()

head(dplyr_database_pull())

dplyr_database_count()

head(dplyr_round_trip())

head(data.table_local())
```

Now let's measure the speeds with `microbenchmark`.

```{r timings}
tm <- microbenchmark(
  "rquery in memory" = nrow(rquery_local()),
  "rquery from db to memory" = nrow(rquery_database_pull()),
  "rquery database count" = rquery_database_count(),
  "rquery database land" = rquery_database_land(),
  "dplyr in memory" = nrow(dplyr_local()),
  "dplyr tbl in memory" = nrow(dplyr_tbl()),
  "dplyr in memory no grouped filter" = nrow(dplyr_local_no_grouped_filter()),
  "dplyr from memory to db and back" = nrow(dplyr_round_trip()),
  "dplyr from db to memory" = nrow(dplyr_database_pull()),
  "dplyr database count" = dplyr_database_count(),
  "dplyr database land" = dplyr_database_land(),
  "data.table in memory" = nrow(data.table_local())
)
saveRDS(tm, "qtimings.RDS")
print(tm)
autoplot(tm)
```

`rquery` appears to be fast.  The extra time for "`rquery` local" is because `rquery`
doesn't *really* have a local mode, it has to copy the data to the database and back
in that case.  I currently guess `rquery` and `dplyr` are both picking up parallelism 
in the database.


```{r sessioninfo}
sessionInfo()
```

```{r dbcleanup}
winvector_temp_db_handle <- NULL
DBI::dbDisconnect(db)
```


